---
title: "Immcantation - enchantR"
subtitle: "RAbHIT haplotype inference"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr RAbHIT haplotype inference"
output: enchantr::immcantation
params:
   input:
      label: "`input`: Path to AIRR TSV file"
      input: file
      value: "input.tsv"
   imgt_db:
      label: "`imgt_db`: Path to germline database"
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   species:
      label: "`species`"
      input: text
      value: "human"
   chain:
      label: "`chain`: IG/TR chain to haplotype"
      input: text
      value: "IGH"
   toHap_col:
      label: "`toHap_col`: comma-separated call columns to haplotype"
      input: text
      value: "v_call,d_call"
   hapBy_col:
      label: "`hapBy_col`: anchor gene call column"
      input: text
      value: "j_call"
   hapBy:
      label: "`hapBy`: anchor gene"
      input: text
      value: "IGHJ6"
   subject_column:
      label: "`subject_column`: Column name containing subject identifiers"
      input: text
      value: "subject_id"
   asc_annotation:
      label: "`asc_annotation`: Convert V calls to PIgLET allele similarity clusters"
      input: checkbox
      value: FALSE
   allele_cluster_table:
      label: "`allele_cluster_table`: Optional ASC table"
      input: file
      value: ""
   min_minor_fraction:
      label: "`min_minor_fraction`: minimum minor anchor allele fraction"
      input: number
      value: 0.3
   kThreshDel:
      label: "`kThreshDel`: lK threshold for deletion calls"
      input: number
      value: 3
   rmPseudo:
      label: "`rmPseudo`: remove pseudo genes"
      input: checkbox
      value: TRUE
   single_gene:
      label: "`single_gene`: restrict haplotyped genes to single-gene calls"
      input: checkbox
      value: TRUE
   run_deletions:
      label: "`run_deletions`: Run double-chromosome deletion support before haplotyping"
      input: checkbox
      value: FALSE
   outname:
      label: "`outname`"
      input: text
      value: "rabhit"
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: "`outdir`: Output directory"
      input: text
      value: !r file.path(getwd(),'enchantr')
   date:
      label: "`date`: Run date"
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo:
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options:
  chunk_output_type: console
---

<div class="introduction_paragraph">
RAbHIT infers immunoglobulin and T-cell receptor haplotypes by using a heterozygous anchor gene to phase other V, D or J genes onto the two chromosomes. For each haplotyped gene, the method compares allele usage across the two anchor-gene alleles and assigns the observed alleles to one or both chromosomes.
</div>

<div class="introduction_paragraph">
The anchor gene must have exactly two well-supported alleles. The `min_minor_fraction` parameter controls the minimum fraction required for the less frequent anchor allele. Genes with weak support can remain unknown, and high Bayes-factor values provide stronger support for chromosome-specific assignments.
</div>

```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 4, fig.path = "figures/",
  echo = params$echo, cache = params$cache,
  warning = FALSE, message = FALSE,
  eval.after = "fig.cap",
  out_dir = params$outdir,
  eval.opts = c("eval", "echo", "fig.height", "fig.width")
)

if (!requireNamespace("rabhit", quietly = TRUE)) {
  stop("The RAbHIT haplotype report requires the 'rabhit' package.")
}

suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("rabhit"))

if (!dir.exists(params$outdir)) {
  dir.create(params$outdir, recursive = TRUE)
}

file.copy(params$logo,
  file.path(params$outdir, "assets", "logo.png"),
  recursive = TRUE, overwrite = TRUE
)

parse_columns <- function(x) {
  cols <- trimws(unlist(strsplit(x, ",", fixed = TRUE)))
  cols[nzchar(cols)]
}

segment_from_call_col <- function(call_col) {
  first <- tolower(substr(call_col, 1, 1))
  seg <- c(v = "V", d = "D", j = "J")[[first]]
  if (is.null(seg)) {
    stop("Cannot infer segment from call column '", call_col, "'. ",
         "Use column names beginning with v, d or j.")
  }
  seg
}

collect_germline <- function(references, chain, call_cols) {
  if (is.null(references[[chain]])) {
    stop("No germline references found for chain '", chain, "'.")
  }
  germline <- unlist(lapply(call_cols, function(call_col) {
    seg <- segment_from_call_col(call_col)
    refs <- references[[chain]][[seg]]
    if (is.null(refs)) {
      stop("No ", seg, " germline references found for chain '", chain, "'.")
    }
    refs
  }))
  germline[!is.na(germline) & nzchar(germline)]
}

toHap_cols <- parse_columns(params$toHap_col)
hapBy_col <- params$hapBy_col
chain <- params$chain
asc_annotation <- as.logical(params$asc_annotation)
run_deletions <- as.logical(params$run_deletions)
rmPseudo <- as.logical(params$rmPseudo)
single_gene <- as.logical(params$single_gene)
genes_order <- NULL
deleted_genes <- c()
nonReliable_Vgenes <- c()
has_haplotype <- FALSE
has_deletions <- FALSE
```

# Input parameters

```{r input-parameters, results='asis'}
printParams(params)
```
```{r echo=FALSE, results='asis'}
print_table_caption(tag = "input", text = "Input parameters.")
save(params, file = file.path(params$outdir, "params.RData"))
```

# Read repertoire data

Load repertoire annotation tables in AIRR format. The report requires the anchor call column (`hapBy_col`), the call columns to be haplotyped (`toHap_col`), and a subject identifier column. Only productive sequences are retained before haplotype inference.

```{r read-input}
cols_to_read <- unique(c(
  "sequence_id", "productive", params$subject_column,
  toHap_cols, hapBy_col, "locus", "species"
))
cols_to_read <- cols_to_read[!is.null(cols_to_read) & nzchar(cols_to_read)]

db <- readInput(params[["input"]], col_select = cols_to_read)

missing_cols <- setdiff(c(params$subject_column, toHap_cols, hapBy_col), colnames(db))
if (length(missing_cols) > 0) {
  stop("Input is missing required columns: ", paste(missing_cols, collapse = ", "))
}

if (params$species == "auto") {
  if ("species" %in% colnames(db)) {
    species <- unique(db[["species"]])
    if (length(species) > 1) {
      stop("Multiple species detected. Only one allowed.")
    }
  } else {
    stop("Can't detect species. The column `species` does not exist in `db`.")
  }
} else {
  species <- params$species
}

if ("productive" %in% colnames(db)) {
  db <- db %>% dplyr::filter(.data[["productive"]] == TRUE)
}

db <- db %>%
  mutate(subject = as.character(.data[[params$subject_column]])) %>%
  filter(!is.na(.data[[hapBy_col]]), .data[[hapBy_col]] != "")

for (call_col in toHap_cols) {
  db <- db %>% filter(!is.na(.data[[call_col]]), .data[[call_col]] != "")
}

if (nrow(db) == 0) {
  stop("No productive records remained after filtering required call columns.")
}

subject_levels <- str_sort(unique(db$subject), numeric = TRUE)
db <- db %>%
  mutate(subject = factor(subject, levels = subject_levels)) %>%
  arrange(subject)

if (!is.null(params$outname) && nzchar(params$outname)) {
  outname <- basename(params$outname)
} else {
  outname <- sub(".tsv(\\.gz)?$", "", basename(params$input))
}

haplotype_output_fn <- paste0(outname, "_haplotype.tsv")
deletion_output_fn <- paste0(outname, "_deletions.tsv")

if (!is.null(params$log)) {
  log_fn <- file.path(params$outdir, paste0(params$log, ".txt"))
} else {
  log_fn <- file.path(params$outdir, paste0(outname, "_command_log.txt"))
}
```

### **This analysis is produced for subject `r paste(unique(db$subject), collapse=", ")` using anchor gene `r params$hapBy`.**

# Read germline database

RAbHIT needs named germline sequences for the genes being haplotyped. The report reads an IMGT-style database and extracts the V, D or J references corresponding to the selected `toHap_col` columns. In ASC mode, V calls and V germlines are converted to PIgLET allele similarity cluster names before haplotyping.

```{r read-germline}
if (is.null(params$imgt_db) || !nzchar(params$imgt_db)) {
  stop("`imgt_db` is required for this report.")
}

imgt_db <- prepareIMGT(params$imgt_db)
references <- dowser::readIMGT(file.path(imgt_db, species, "vdj"), quiet = TRUE)
toHap_GERM <- collect_germline(references, chain, toHap_cols)

if (length(toHap_GERM) == 0) {
  stop("No germline sequences were found for the requested haplotype columns.")
}
```

# Allele similarity clusters (optional)

When enabled, ASC conversion maps V allele calls and V germlines to PIgLET allele similarity cluster names before haplotype inference. This can reduce ambiguity among highly similar or duplicated IGHV alleles while preserving the RAbHIT anchor-based haplotyping model.

```{r asc-conversion, eval=asc_annotation}
if (asc_annotation) {
  v_cols <- toHap_cols[tolower(substr(toHap_cols, 1, 1)) == "v"]
  if (length(v_cols) != 1) {
    stop("ASC conversion currently requires exactly one V call column in `toHap_col`.")
  }

  allele_cluster_table <- NULL
  if (!is.null(params$allele_cluster_table) &&
      nzchar(params$allele_cluster_table) &&
      params$allele_cluster_table != "NULL") {
    allele_cluster_table <- read.delim(params$allele_cluster_table, stringsAsFactors = FALSE)
  }

  v_germline <- references[[chain]][["V"]]
  asc <- rabhit::convertToASC(
    clip_db = db,
    germline = v_germline,
    chain = chain,
    allele_cluster_table = allele_cluster_table,
    v_call = v_cols[[1]]
  )

  db <- asc$clip_db
  genes_order <- asc$genes_order
  non_v_cols <- setdiff(toHap_cols, v_cols)
  non_v_germline <- if (length(non_v_cols) > 0) collect_germline(references, chain, non_v_cols) else NULL
  toHap_GERM <- c(asc$germline, non_v_germline)
}
```

# Deletion support (optional)

RAbHIT can use prior evidence for double-chromosome deletions and non-reliable V genes. If enabled, the report estimates these annotations before haplotyping and passes them to `createFullHaplotype()`.

```{r deletion-support, eval=run_deletions}
if (run_deletions) {
  nonReliable_Vgenes <- rabhit::nonReliableVGenes(db, chain = chain)
  deleted_genes <- rabhit::deletionsByBinom(
    db,
    chain = chain,
    nonReliable_Vgenes = nonReliable_Vgenes
  )
  has_deletions <- is.data.frame(deleted_genes) && nrow(deleted_genes) > 0
}
```

# RAbHIT haplotype inference

`createFullHaplotype()` phases the requested genes by comparing their allele counts across the two alleles of the anchor gene. The output contains one row per haplotyped gene and subject, with chromosome-specific assignments named by the two anchor alleles.

The `kThreshDel` parameter controls the Bayes-factor threshold used to call single-chromosome deletions. The `single_gene` option restricts haplotyping to single-gene assignments, reducing ambiguity from multi-gene calls.

```{r rabhit-haplotype-inference}
haplo_db <- rabhit::createFullHaplotype(
  clip_db = db,
  toHap_col = toHap_cols,
  hapBy_col = hapBy_col,
  hapBy = params$hapBy,
  toHap_GERM = toHap_GERM,
  relative_freq_priors = TRUE,
  kThreshDel = as.numeric(params$kThreshDel),
  rmPseudo = rmPseudo,
  deleted_genes = deleted_genes,
  nonReliable_Vgenes = nonReliable_Vgenes,
  min_minor_fraction = as.numeric(params$min_minor_fraction),
  single_gene = single_gene,
  chain = chain
)

if (is.null(haplo_db) || nrow(haplo_db) == 0) {
  stop("RAbHIT did not return haplotype rows. Check anchor gene heterozygosity and min_minor_fraction.")
}

has_haplotype <- TRUE
```

# Haplotype inference results

## Haplotype inference table

<div class="introduction_paragraph">
The haplotype table summarizes chromosome-specific allele assignments for each gene. The two columns immediately after `gene` correspond to the two anchor-gene alleles and represent the inferred alleles on each chromosome.
</div>

Interpret the columns as follows:

- `subject`: subject identifier used by RAbHIT.
- `gene`: haplotyped gene.
- anchor allele columns: inferred chromosome-specific assignments.
- `alleles`: candidate alleles observed for the gene.
- `counts1`-`counts4`: allele counts split across anchor chromosomes.
- `k1`-`k4`: Bayes-factor support for the corresponding allele assignment.
- `Del`, `NR` or `Unk` values indicate deletion, non-reliable annotation, or unknown assignment when present.

Reliability notes:

- The anchor gene must have exactly two alleles after filtering.
- Lower minor-anchor fractions reduce haplotyping reliability and may prevent inference.
- Larger `k` values indicate stronger support for chromosome-specific assignment.
- Multi-gene assignments can inflate ambiguity; `single_gene = TRUE` restricts inference to single-gene calls.

```{r haplotype-table, results='asis'}
caption <- "RAbHIT haplotype inference results showing chromosome-specific allele assignments."

tab_haplotype <- eetable(haplo_db,
  caption = caption,
  outdir = params$outdir,
  file = "tab_haplotype"
)
tab_haplotype$table
```

```{r echo=FALSE, results='asis'}
print_table_caption(tag = "tab_haplotype", text = tab_haplotype$caption)
```

## Haplotype map

<div class="introduction_paragraph">
The haplotype map displays each gene along the chromosome, its allele assignment on each anchor-defined chromosome, and the assignment confidence. Unknown, deleted, or non-reliable calls should be interpreted together with the count and Bayes-factor columns in the table.
</div>

```{r haplotype-map, results='asis', echo=FALSE}
subjects <- unique(as.character(haplo_db$subject))

for (i in seq_along(subjects)) {
  plot_env <- new.env()
  plot_env$subject_id <- subjects[[i]]
  plot_env$haplo_subject <- haplo_db[as.character(haplo_db$subject) == plot_env$subject_id, , drop = FALSE]
  plot_env$params <- params
  plot_env$chain <- chain
  plot_env$genes_order <- genes_order
  plot_env$caption <- paste0("RAbHIT haplotype map for subject ", plot_env$subject_id, ".")
  plot_env$fig_height <- max(6, min(18, nrow(plot_env$haplo_subject) * 0.28))

  child <- paste0(
    "\n### ", plot_env$subject_id, " \n\n",
    "```{r haplotype-map-", i, ", echo=FALSE, message=FALSE, warning=FALSE, fig.height=", plot_env$fig_height, ", fig.width=12, fig.cap=caption}\n",
    "rabhit::plotHaplotype(haplo_subject, genes_order=genes_order, chain=chain)\n",
    "```\n"
  )
  cat(knitr::knit_child(text = child, envir = plot_env, quiet = TRUE))
}
```

```{r haplotype-heatmap-intro, eval=length(unique(as.character(haplo_db$subject))) > 1, results='asis', echo=FALSE}
cat("## Multi-subject haplotype heatmap\n\n")
cat("The heatmap compares chromosome-specific allele assignments across subjects. RAbHIT returns suggested plot dimensions based on the number of samples and genes shown.\n\n")
```

```{r haplotype-heatmap, eval=length(unique(as.character(haplo_db$subject))) > 1, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
hap_heatmap <- rabhit::hapHeatmap(haplo_db, chain = chain, genes_order = genes_order)
replayPlot(hap_heatmap$p)
```

```{r haplotype-dendrogram-intro, eval=length(unique(as.character(haplo_db$subject))) > 1 && requireNamespace("ggdendro", quietly = TRUE), results='asis', echo=FALSE}
cat("## Haplotype dendrogram\n\n")
cat("The dendrogram clusters subjects by haplotype similarity using RAbHIT's Jaccard-distance comparison of chromosome-specific assignments.\n\n")
```

```{r haplotype-dendrogram, eval=length(unique(as.character(haplo_db$subject))) > 1 && requireNamespace("ggdendro", quietly = TRUE), echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
rabhit::hapDendo(haplo_db, chain = chain, genes_order = genes_order)
```

```{r deletion-table, eval=has_deletions, results='asis'}
caption <- "RAbHIT deletion support table generated before haplotype inference."

tab_deletions <- eetable(deleted_genes,
  caption = caption,
  outdir = params$outdir,
  file = "tab_deletions"
)
tab_deletions$table
```

```{r echo=FALSE, results='asis', eval=has_deletions}
print_table_caption(tag = "tab_deletions", text = tab_deletions$caption)
```

```{r deletion-plot, eval=has_deletions, echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=8}
rabhit::plotDeletionsByBinom(deleted_genes)
```

# Save haplotype report

```{r save-haplotype}
haplotype_output_dir <- file.path(params$outdir, "haplotypes")
dir.create(haplotype_output_dir, recursive = TRUE, showWarnings = FALSE)

write.table(
  haplo_db,
  file = file.path(haplotype_output_dir, haplotype_output_fn),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

if (has_deletions) {
  write.table(
    deleted_genes,
    file = file.path(haplotype_output_dir, deletion_output_fn),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}
```

```{r log-end}
cat("START> RAbHIT Haplotype Inference", file = log_fn, append = FALSE)
cat(paste0("\nSUBJECT> ", paste(unique(db$subject), collapse = ", ")), file = log_fn, append = TRUE)
cat(paste0("\nMETHOD> RAbHIT anchor haplotype"), file = log_fn, append = TRUE)
cat(paste0("\nFILE> ", basename(params$input)), file = log_fn, append = TRUE)
cat(paste0("\nCHAIN> ", chain), file = log_fn, append = TRUE)
cat(paste0("\nANCHOR_COLUMN> ", hapBy_col), file = log_fn, append = TRUE)
cat(paste0("\nANCHOR_GENE> ", params$hapBy), file = log_fn, append = TRUE)
cat(paste0("\nTO_HAP_COLUMNS> ", paste(toHap_cols, collapse = ",")), file = log_fn, append = TRUE)
cat(paste0("\nASC_ANNOTATION> ", asc_annotation), file = log_fn, append = TRUE)
cat(paste0("\nRUN_DELETIONS> ", run_deletions), file = log_fn, append = TRUE)
cat(paste0("\nOUTPUT> ", file.path(haplotype_output_dir, haplotype_output_fn)), file = log_fn, append = TRUE)
if (has_deletions) {
  cat(paste0("\nOUTPUT_DELETIONS> ", file.path(haplotype_output_dir, deletion_output_fn)), file = log_fn, append = TRUE)
}
cat(paste0("\nHAPLOTYPE_ROWS> ", nrow(haplo_db)), file = log_fn, append = TRUE)
```

```{r, child=c('versions.Rmd')}
```
