--- 
title: "Immcantation - enchantR"
subtitle: "Single Cell Quality Control"
author: "`r params$author`"
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantR Single Cell QC Report"
output: enchantr::immcantation
params:
   author:
       value: "Authors: Susanna Marquez, Edel Aron"
   input: 
      label: "`input`: Path to repertoires file"
      input: file
      value: "input.tsv"
   outname:
      label: "`outname`"
      input: text
      value: "scqc"
   outputby:
      label: "`outputby` name of the column in `input` that contains identifiers that will be used to split the output db."      
      input: text
      value: "sample_id"      
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: '`outdir`: Output directory'
      input: text
      value: !r file.path(getwd(),'enchantr')
   date: 
      label: '`date`: Run date'
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo: 
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options: 
  chunk_output_type: console
---

```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4, fig.path = "figures/",
                      echo = params$echo, cache = params$cache,
                      warning = FALSE, message = FALSE,
                      eval.after = "fig.cap", out_dir = params$outdir,
                      eval.opts = c("eval", "echo", "fig.height", "fig.width"))

# libraries
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("alakazam"))
suppressPackageStartupMessages(library("shazam"))
suppressPackageStartupMessages(library("plotly"))

if (!dir.exists(params[["outdir"]])) {
  dir.create(params[["outdir"]], recursive = TRUE)
}

file.copy(params$logo,
          file.path(params$outdir, "assets", "logo.png"),
          recursive = TRUE, overwrite = TRUE)
```

# Input parameters

```{r input-parameters, results="asis", echo=FALSE}
printParams(params)
```

```{r echo=FALSE, results="asis", echo=FALSE}
print_table_caption(tag = "input", text = "Input parameters.")
save(params, file = file.path(params$outdir, "params.RData"))
```

# Read repertoires

```{r, results="asis"}
# Read repertoire(s)
db <- readInput(params[["input"]], col_select = NULL)

input_sizes <- db %>%
                 group_by(!!!rlang::syms(unique(c("input_file")))) %>%
                 summarize(input_size = n())

# Track input size of the expected output groups, for the
# end of report summary
input_sizes_byoutput <- db %>%
                 group_by(!!!rlang::syms(unique(params$outputby))) %>%
                 summarize(input_size = n(),
                           num_input_files = length(unique(input_file)),
                           input_file = paste(unique(input_file),
                                              collapse = ","))

input_size <- nrow(db)

input_summary <- db %>%
                   group_by(!!!rlang::syms(unique(c("sample_id",
                                                    params$outputby,
                                                    "locus")))) %>%
                   summarize(n = n()) %>%
                   pivot_wider(id_cols = sample_id,
                               values_from = n, names_from = locus)

caption_text_rep <- ifelse(nrow(input_summary) > 1, "repertoires", "repertoire")
caption_text <- paste("Summary of input data. Displays the number of sequences per sample and 
                       per locus. There was", nrow(input_summary), caption_text_rep,
                      "loaded with", input_size, "sequences in total.")

eetable(input_summary, caption = caption_text)$table
```

```{r echo=FALSE, results="asis"}
print_table_caption(tag = "input_summary", text = caption_text)
```

# Sequence distribution

Distribution of the number of sequences per cell and locus

```{r, results="asis"}
seqs_per_cell <- countSequencesPerCell(db)
```

```{r p-seqs-per-cell, fig.width=7, fig.height=min(50,length(unique(seqs_per_cell[["sample_id"]]))*2), fig.cap=p_seqs_per_cell$enchantr$html_caption}
p_seqs_per_cell <- plotSequencesPerCell(seqs_per_cell) + theme_enchantr()
p_seqs_per_cell <- eeplot(p_seqs_per_cell, params$outdir,
                          knitr::opts_current$get("label"))
ggplotly(p_seqs_per_cell)
```

```{r, results="asis"}
tab <- eetable(seqs_per_cell %>% dplyr::filter(cell_num_sequences > 1),
               caption = "Cells with more than one sequence")
tab$table
```

```{r echo=FALSE, results="asis"}
print_table_caption(tag = "seqs_per_cell", text = tab$caption)
```

# Filtration

## Keep only productive cells

## Remove cells without heavy chains

Since most of the analyses are based on heavy chains, we remove cells with only light chains.

```{r findLightOnlyCells, eval=nrow(db)>0, results="asis"}
db <- findLightOnlyCells(db, sample_id = "sample_id", cell_id = "cell_id",
                         locus = "locus", fields = NULL)
light_only_size <- sum(db[["light_only_cell"]])

light_only <- db %>%
                rowwise() %>%
                mutate(light_only_cell = if (light_only_cell == TRUE) {"light_only"}
                       else {"has_heavy"}) %>%
                select(sample_id, cell_id, light_only_cell) %>%
                distinct() %>%
                group_by(sample_id, light_only_cell) %>%
                summarize(num_cells = n(), .groups = "drop") %>%
                pivot_wider(names_from = light_only_cell,
                            values_from = num_cells)

# Remove the cells with only light chains
db <- db %>% dplyr::filter(!light_only_cell)
db_with_heavy_size <- nrow(db)

caption_text_seq <- ifelse(light_only_size > 0, "sequences", "sequence")
tab <- eetable(light_only,
               caption = paste("Cells with only light chains account for",
                               light_only_size, caption_text_seq, "removed."))
tab$table

if(nrow(db)==0){
  cat("\nNo cells left after removing cells without heavy chains.\n\n")
}
```

```{r echo=FALSE, results="asis"}
print_table_caption(tag = "light_only", text = tab$caption)
```

## Remove cells with multiple heavy chains

When calling clones (B cells that descend from a common naive B cell ancestor) from single cell data, SCOPer will throw an error message such as “xxx cell(s) with multiple heavy chains found. One heavy chain per cell is expected” if any cells in the data contain multiple heavy chains and then stop running.

A simple solution is just to remove cells with multiple heavy chains from the single cell data:

```{r removeDoublets, eval=nrow(db)>0, results = 'asis'}
db <- removeDoublets(db, cell_id = "cell_id", locus = "locus",
                     sequence_id = "sequence_id", fields = "sample_id")
db_no_doublets_size <- nrow(db)

cat(paste(db_with_heavy_size-db_no_doublets_size, "sequences have been removed because they were found in cells with multiple heavy chains."))

if(nrow(db)==0){
  cat("No cells left after removing cells with multiple heavy chains.\n\n")
}

```

# Check for contamination

```{r}
# check whether to check contamination
sample_count <- n_distinct(db$sample_id)
calculate_contamination <- sample_count > 1
dups_found <- FALSE
```

Sequences in different samples that share the same `cell_id` and nucleotide sequence (`sequence_alignment`) can indicate contamination.

```{r sc-duplicates, eval=calculate_contamination, results="asis"}
# Find duplicated sequences
dups <- findSingleCellDuplicates(db, fields = "sample_id", cell_id = "cell_id",
                                 seq = "sequence_alignment", mode = "sequences")
dups_found <- sum(dups$dups$sc_duplicate, na.rm = T) > 0
```

```{r sc-duplicates-zero, results="asis", eval=!dups_found, echo=FALSE}
cat("No suspicious overlaps detected.\n\n")
```

```{r sc-duplicates-summary, results="asis", eval=dups_found}
dups$dups <- dups$dups %>%
               group_by(sample_id, cell_id) %>%
               mutate(sc_duplicate_cell = if_else(any(sc_duplicate),
                                                  TRUE, FALSE))

num_cells_with_dups <- dups$dups %>%
                         dplyr::filter(sc_duplicate_cell) %>%
                         select(sample_id, cell_id) %>%
                         distinct() %>%
                         nrow()

cat(paste(sum(dups$dups$sc_duplicate), "duplicated sequences detected. Removing ",
          sum(dups$dups$sc_duplicate_cell), "sequences belonging to",
          num_cells_with_dups , "suspicious cells."))

dups_summary <- dups[["dups"]] %>%
    group_by(sample_id) %>%
    summarize(duplicates = sum(sc_duplicate_cell, na.rm = TRUE),
              not_duplicates = sum(!sc_duplicate_cell, na.rm = TRUE)
              ) %>%
    rowwise() %>%
    mutate(duplication_percent = 100 * duplicates /
               sum(duplicates + not_duplicates))

tab_dups_summary <- eetable(
    dups_summary %>% mutate(duplication_percent = round(duplication_percent, 2)),
    caption = "Summary of duplicated sequences: sequences belonging to cells
               with suspicious sequences (sequences that have the same cell_id
               and sequence across samples). The percent represents the number
               of duplicated sequences over the total of sequences in the
               sample.")

tab_dups_summary$table %>% 
    formatStyle("duplication_percent", target = "cell",
                backgroundColor =
                  styleInterval(0:100, round(seq(255, 40,
                                                 length.out = length(0:100) + 1), 0) %>%
                                  {paste0("rgb(255,", ., ",", ., ")")})
    )
```

```{r echo=FALSE, results="asis", eval=dups_found}
print_table_caption(tag = "dups_summary", text = tab_dups_summary$caption)
```

## Stacked bar plot

```{r p-sc-duplicates, fig.cap=p_sc_duplicates$enchantr$html_caption, results="asis", eval=dups_found}
caption <- "Number of sequences that share cell_id and sequence with other samples."
p_sc_duplicates <- ggplot(dups[["dups"]],
                          aes(x = sample_id, fill = sc_duplicate_cell)) +
                     geom_bar(linewidth = 0.2, color = "black") +
                     coord_flip() +
                     theme_enchantr()
p_sc_duplicates <- eeplot(p_sc_duplicates, params$outdir,
                          knitr::opts_current$get("label"),
                          caption = caption)

ggplotly(p_sc_duplicates) %>% layout(barmode = "stack")
```

## Cell overlaps matrix

```{r overlap-summary, eval=dups_found}
# find the number of cells having same seq and cell barcode
dup_count <- singleCellSharingCounts(dups)

if (nrow(dup_count) > 0) {
    overlap_summary <- dup_count %>%
        filter(overlap_percent > 0 ) %>%
        filter(sc_duplicate_group != name)
    
    overlap_summary_tab <- eetable(
        overlap_summary,
        caption = "Summary of duplicated cells: cells that have sequences with a
                   combination of cell_id and sequence found in multiple
                   samples."
    )
    
    overlap_summary_tab$table %>% 
        formatStyle("overlap_percent", target = "cell",
                    backgroundColor =
                      styleInterval(0:100, round(seq(255, 40,
                                                     length.out = length(0:100) + 1), 0) %>%
                                      {paste0("rgb(255,", ., ",", ., ")")})
        )
}
```

```{r echo=FALSE, results="asis", eval=dups_found }
print_table_caption(tag = "overlap_summary", text = overlap_summary_tab$caption)
```

```{r p-overlap-matrix, eval=TRUE, fig.cap=caption, eval=dups_found}
# plot duplicated cells
caption <- NULL
if (nrow(dup_count) > 0) {
  p_overlap_matrix <- plotOverlapSingleCell(dup_count) + theme_enchantr()
  p_overlap_matrix <- eeplot(p_overlap_matrix, params$outdir,
                             knitr::opts_current$get("label"),
                             caption = "Duplicated cells.")
  caption <- p_overlap_matrix$enchantr$html_caption
  
  ggplotly(p_overlap_matrix)
}
```

# Make cell_id and sequence_id unique

There could still exist duplicated `cell_id` or `sequence_id`,
that are not doublets or suspected contaminants, between samples. This
could cause issues downstream. For example, when finding clones by subject, if
different samples of the same subject happen to share `cell_id`, 
`scoper::hierarchicalClones` would stop with an error.

To avoid this, we add a numeric suffix (`-1`, `-2`,... up to the number of 
samples) to `sequence_id` and `cell_id`.

# Final repertoires 

```{r outsetup, echo=FALSE}
# Remove cells (all sequences in cells) with duplicated sequences
# pass <- qc_results$log$scqc_pass
if (dups_found) {
  pass <- dups[["dups"]][["sc_duplicate_cell"]] == FALSE
} else {
  pass <- rep(TRUE, nrow(db))
}

if (!is.null(params$outname)) {
  output_fn <- paste0(params$outname, "_scqc-pass.tsv")
} else {
  output_fn <- sub(".tsv$", "_scqc-pass.tsv", basename(params$input))
}

if (!is.null(params$log)) {
  log_fn <- paste0(params$log, ".txt")
} else {
  log_fn <- sub("_scqc-pass.tsv$", "_command_log.txt", basename(output_fn))
}
```

```{r dbpass, results="asis", echo=FALSE}
# Save data, sequences that pass, while making cell_id and sequence id unique
db <- db[pass, , drop=FALSE]

# Global unique cell_id
sample_ids <- unique(db[["sample_id"]])
db <- db %>%
        mutate(cell_id = paste(cell_id, match(sample_id, sample_ids),
                               sep = "-"),
               sequence_id = paste(sequence_id, match(sample_id, sample_ids),
                                   sep = "-"))

output_groups <- db %>%
                   ungroup() %>%
                   select(!!!rlang::syms(c(params$outputby, "input_file"))) %>%
                   group_by(!!!rlang::syms(params$outputby)) %>%
                   summarize(output_size = n(),
                             num_input_files = length(unique(input_file)),
                             input_file = paste(unique(input_file), collapse=",")) %>%
                   right_join(input_sizes_byoutput) %>%
                   select(sample_id, input_size, output_size, everything())
output_groups[is.na(output_groups)] <- 0

output_summary_tab <- eetable(
    output_groups,
    caption = paste0("Output summary.")
)
output_summary_tab$table
```

```{r output_summary, results="asis", echo=FALSE}
print_table_caption(tag = "output_summary_tab", text = output_summary_tab$caption)
```

Find your final repertoires here, saved by `r paste(params$outputby, collapse=", ")`.

```{r dbpass-save, results="asis", echo=FALSE}
if (any(output_groups[["num_input_files"]] > 1)) {
  stop("Unhandled.")
}

output_files <- c()
output_sizes <- c()
dir.create(file.path(params$outdir, "repertoires"))
for (i in 1:nrow(output_groups)) {
    this_group <- output_groups[i, , drop = F]
    group_fn <- ""
    group_db <- data.frame()
    
    if (this_group[["output_size"]]>0) {
        group_db <- right_join(db, this_group,
                               by = c(params$outputby, "input_file")) %>%
            select(-input_file, -num_input_files, -input_size, -output_size)
        
        group_db_label <- gsub("^id_", "",
                               makeLabel(group_db, fields = params$outputby))
        group_fn <- file.path(params$outdir, "repertoires",
                              paste0(group_db_label, "_", output_fn))
        
        write_rearrangement(group_db, file = group_fn)
        cat("- [", group_db_label, 
            "](", file.path("repertoires", basename(group_fn)), ")\n", sep = "")
    }
    
    output_files <- c(output_files, group_fn)
    output_sizes <- c(output_sizes, nrow(group_db))
}
```

```{r log, echo=FALSE}
cat("START> SingleCellQC", file = log_fn, append = FALSE)

for (i in 1:nrow(output_groups)) {
  cat(paste0("\nFILE", i, "> ", output_groups[["input_file"]][i]),
      file = log_fn, append = T)
  cat(paste0("\nOUTPUT", i, "> ", basename(output_files[i])),
      file = log_fn, append = T)
  cat(paste0("\nPASS", i, "> ", output_sizes[i]),
      file = log_fn, append = T)
  cat(paste0("\nFAIL", i, "> ", output_groups[["input_size"]][i] - output_sizes[i]),
      file = log_fn, append = T)
}
```

```{r, child=c("versions.Rmd")}
```
