---
title: "Immcantation - enchantR"
subtitle: "Genotype inference"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr genotype inference"
output: enchantr::immcantation
params:
   input:
      label: "`input`: Path to AIRR TSV file"
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/refs/heads/airrflow/testdata-genotyping/airr_data_to_test_novel_alleles.tsv"
   imgt_db:
      label: "`imgt_db`: Path to germline FASTA file"
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   novel_allele_evidence:
      label: "`novel_allele_evidence`: Path to novel allele evidence file"
      input: file
      value: NULL
   species:
      label: "`species`"
      input: text
      value: "human"
   segments:
      label: "`segments`: which segments to reassign. Options are 'v', 'd', 'j'. Multiple segments can be provided as a comma-separated string. auto will reassign segments based on loci type."    
      input: text
      value: "auto"
   v_call:
      label: "`v_call`: Column name for V gene calls"
      input: text
      value: "v_call"
   j_call:
      label: "`j_call`: Column name for J gene calls"
      input: text
      value: "j_call"
   d_call:
      label: "`d_call`: Column name for D gene calls"
      input: text
      value: "d_call"
   seq_column:
      label: "`seq_column`: Column name of segment sequence (e.g., sequence_alignment)"
      input: text
      value: "sequence_alignment"
   germline_column:
      label: "`germline_column`: Column name of germline gene call (e.g., germline_alignment)"
      input: text
      value: "germline_alignment"
   clone_id_column:
      label: "`clone_id_column`: Column name of clone ID (e.g., clone_id)"
      input: text
      value: "clone_id"
   single_clone_representative:
      label: "`single_clone_representative`: Use single representative per clone"
      input: checkbox
      value: FALSE
   find_unmutated:
      label: "`find_unmutated`: Find unmutated sequences for genotype inference"
      input: checkbox
      value: FALSE
   single_assignments:
      label: "`single_assignments`: Filter to single gene assignments only. Keep FALSE for IMGT references with duplicate alleles (e.g. IGHV1-69D*01), which only appear in ambiguous multi-allele calls and would otherwise never be genotyped."
      input: checkbox
      value: FALSE
   method:
      label: "`method`: Genotype inference method. One of 'bayesian' (TIgGER Bayesian), 'fraction' (TIgGER cumulative fraction), or 'allele_based' (PIgLET allele-specific thresholds)."
      input: select
      choices: ["bayesian", "fraction", "allele_based"]
      value: "bayesian"
   v_priors:
      label: "`v_priors`: [method='bayesian'] V gene priors (comma-separated)"
      input: text
      value: "0.6,0.4,0.4,0.35,0.25,0.25,0.25,0.25,0.25"
   d_priors:
      label: "`d_priors`: [method='bayesian'] D gene priors (comma-separated)"
      input: text
      value: "0.5,0.5,0,0,0,0,0,0,0"
   j_priors:
      label: "`j_priors`: [method='bayesian'] J gene priors (comma-separated)"
      input: text
      value: "0.5,0.5,0,0,0,0,0,0,0"
   fraction_to_explain:
      label: "`fraction_to_explain`: [method='fraction'] Cumulative fraction of a gene's sequences that the genotyped alleles must explain."
      input: numeric
      value: 0.875
   gene_cutoff:
      label: "`gene_cutoff`: [method='fraction'] Alleles below this frequency are discarded as noise."
      input: numeric
      value: 0.0001
   allele_thresholds_db:
      label: "`allele_thresholds_db`: [method='allele_based'] Path to a TSV of alleles and their population-derived presence thresholds. If NULL, PIgLET's bundled thresholds are used."
      input: file
      value: NULL
   allele_thresholds_release:
      label: "`allele_thresholds_release`: [method='allele_based'] Release or version label for the threshold table, recorded with the genotype. The checksum is recorded either way."
      input: text
      value: ""
   default_threshold:
      label: "`default_threshold`: [method='allele_based'] Threshold used for alleles absent from the threshold table."
      input: numeric
      value: 0.0001
   asc_annotation:
      label: "`asc_annotation`: [method='allele_based'] Are the allele calls annotated with allele similarity clusters?"
      input: checkbox
      value: FALSE
   translate_to_asc:
      label: "`translate_to_asc`: [method='allele_based'] Collapse identical alleles into allele similarity clusters for inference."
      input: checkbox
      value: FALSE
   depth_adjusted_threshold:
      label: "`depth_adjusted_threshold`: [method='allele_based'] Adjust allele thresholds based on the depth of the repertoire."
      input: checkbox
      value: TRUE
   z_score_threshold:
      label: "`z_score_threshold`: [method='allele_based'] Z-score threshold filtering for allele-based inference."
      input: numeric
      value: 0
   genotypeby:
      label: "`genotypeby` name of the column in `input` that contains sample identifiers that will be used to genotype the input db." 
      input: text
      value: NULL
   outname:
      label: "`outname`"
      input: text
      value: "sample"
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: "`outdir`: Output directory"
      input: text
      value: !r file.path(getwd(),'enchantr')
   date:
      label: "`date`: Run date"
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo:
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options:
  chunk_output_type: console
---

<div class="introduction_paragraph">
Genotype inference determines the set of germline alleles carried by an individual for each immunoglobulin (IG) or T-cell receptor (TR) gene. 
Because repertoire sequencing captures rearranged V(D)J gene assignments rather than the germline loci themselves, an individual's true germline composition must be inferred from the distribution of unmutated or minimally mutated sequences observed across the repertoire. 
Accurate genotype inference corrects for missing or misassigned alleles in the reference database, ensuring that subsequent analyses—such as mutation quantification, lineage reconstruction, and repertoire comparison—are based on the subject’s actual germline profile.
</div>

<div class="introduction_paragraph">
<strong>Note:</strong> This method has not been thoroughly tested for genotyping single-cell repertoires.
</div>

```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 4, fig.path = "figures/",
  echo = params$echo, cache = params$cache,
  warning = FALSE, message = FALSE,
  eval.after = "fig.cap",
  out_dir = params$outdir,
  eval.opts = c("eval", "echo", "fig.height", "fig.width")
)

# Libraries
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("tigger"))

if (!dir.exists(params$outdir)) {
  dir.create(params$outdir, recursive = TRUE)
}

# One child chunk per plot, for per-figure height and caption.
emit_plot_sections <- function(plots, prefix, caption_for, fig_height) {
  fig_height <- rep_len(fig_height, length(plots))
  for (i in seq_along(plots)) {
    plot_env <- new.env()
    plot_env$p_obj <- plots[[i]]
    plot_env$ix <- i
    plot_env$params <- params
    plot_env$caption <- caption_for(names(plots)[i])
    child <- paste0(
      "\n## ", names(plots)[i], " \n\n",
      "```{r ", prefix, "-", i, ", echo=FALSE, message=FALSE, warning=FALSE, fig.height=",
      fig_height[i], ", fig.width=8, fig.cap=caption}\n",
      "p <- eeplot(p_obj, outdir=params$outdir, file=paste0('", prefix, "-', ix), caption=caption)\n",
      "plot(p)\n",
      "caption <- p$enchantr$html_caption\n",
      "```\n"
    )
    cat(knitr::knit_child(text = child, envir = plot_env, quiet = TRUE))
  }
}

# Keep only the parameters the selected method uses.
method <- params$method
report_params <- check_genotype_params(params, method)
method_label <- genotype_method_label(method)

# Convert boolean parameters
find_unmutated <- as.logical(params$find_unmutated)
single_assignments <- as.logical(params$single_assignments)
genotypeby <- params$genotypeby
has_novel <- has_path(params$novel_allele_evidence)
has_novel_table <- FALSE
has_novel_figures <- FALSE

file.copy(params$logo,
  file.path(params$outdir, "assets", "logo.png"),
  recursive = TRUE, overwrite = TRUE
)
```

# Input parameters

```{r input-parameters, results='asis'}
printParams(report_params)
```
```{r echo=FALSE, results='asis'}
print_table_caption(tag = "input", text = "Input parameters.")
save(params, file = file.path(params$outdir, "params.RData"))
```

# Read repertoire data

Load repertoire annotation tables in AIRR format.
Expected columns include `sequence_id`, `productive`, `v_call` / `d_call` / `j_call`,  `sequence_alignment`. Only productive, in-frame sequences without stop codons are retained to avoid artefacts.


```{r read-input}
# Read repertoire data
cols_to_read <- unique(c(
  "sequence_id", "productive", "junction_length", "sequence_alignment",
  params$v_call, params$d_call, params$j_call,
  params$germline_column, params$clone_id_column,
  params$genotypeby, "subject_id", "sample_id", "species"
))

db <- readInput(params[["input"]], col_select = cols_to_read)

# Standardize call column names for downstream helpers, which expect
# v_call/d_call/j_call internally.
for (seg in c("v", "d", "j")) {
  seg_call_param <- params[[paste0(seg, "_call")]]
  seg_call_col <- paste0(seg, "_call")
  if (!is.null(seg_call_param) && seg_call_param %in% names(db) && seg_call_param != seg_call_col) {
    db[[seg_call_col]] <- db[[seg_call_param]]
  }
}

if (params$species == "auto") {
  if ("species" %in% colnames(db)) {
    species <- unique(db[["species"]])
    if (length(species) > 1) {
      stop("Multiple species detected. Only one allowed.")
    }
  } else {
    stop("Can't detect species. The column `species` does not exist in `db`.")
  }
} else {
  species <- params$species
}

if (!"locus" %in% colnames(db)) {
  db[["locus"]] <- getLocus(db[["v_call"]])
}
heavy_chains <- isHeavyChain(db[["locus"]])

# segment to genotype
if (params$segments == "auto") {
  segments_to_genotype <- if (any(heavy_chains)) c("v", "d", "j") else c("v", "j")
} else {
  segments_to_genotype <- unlist(strsplit(params$segments, ","))
  segments_to_genotype <- tolower(trimws(segments_to_genotype))
  valid_segments <- c("v", "d", "j")
  if (!all(segments_to_genotype %in% valid_segments)) {
    stop("Invalid segments provided in `segments` parameter. Valid options are 'v', 'd', 'j'.")
  }
  if (any(segments_to_genotype == "d") && !any(heavy_chains)) {
    stop("D segments can only be reassigned for heavy chain repertoires.")
  }
}


if (all(c("subject_id", "sample_id") %in% colnames(db))) {
  subject_levels <- str_sort(unique(db$subject_id), numeric = TRUE)
  sample_levels <- str_sort(unique(db$sample_id), numeric = TRUE)
  db <- db %>%
    mutate(
      subject_id = factor(subject_id, levels = subject_levels),
      sample_id  = factor(sample_id, levels = sample_levels)
    ) %>%
    arrange(subject_id, sample_id)
} else {
  # add fake columns
  db <- db %>%
    mutate(
      subject_id = "subject",
      sample_id  = "sample"
    )
}

# Each filter records rows in and out in `filter_steps`.
filter_steps <- NULL
rows_in <- nrow(db)
db <- db %>% dplyr::filter(.data[["productive"]] == TRUE)
filter_steps <- count_rows_step(filter_steps, "productive == TRUE", rows_in, nrow(db))

rows_in <- nrow(db)
db <- db %>% dplyr::filter(.data[["junction_length"]] %% 3 == 0)
filter_steps <- count_rows_step(filter_steps, "junction_length %% 3 == 0", rows_in, nrow(db))

if (nrow(db) == 0) {
  stop("No sequences are left after filtering. See the row counts above: ",
       paste(sprintf("%s %d->%d", filter_steps$step, filter_steps$rows_in, filter_steps$rows_out),
             collapse = "; "))
}

db_before_clone_collapse <- db
db_raw_productive <- db_before_clone_collapse

# Set up output filenames
outname <- if (!is.null(params$outname)) basename(params$outname) else sub("\\.tsv(\\.gz)?$", "", basename(params$input))
report_output_fn <- paste0(outname, "_genotype_report.tsv")
log_fn <- file.path(params$outdir, paste0(if (!is.null(params$log)) params$log else "command_log", ".txt"))
```


# Read germline database

Import the nucleotide germline reference used for alignment.
Using the same reference ensures positional numbering and mutation counts remain consistent with the upstream alignment step. Each allele call present in the repertoire should have a corresponding entry in the germline database.


```{r read-germline}
loci <- unique(db[["locus"]])

# Handle germline database input
if (params$imgt_db != "") {
  imgt_db <- prepareIMGT(params$imgt_db)
  references <- dowser::readIMGT(file.path(imgt_db, species, "vdj"), quiet = TRUE)
} else {
  references <- NA
}
```

# Clonal representative selection (optional)

When multiple sequences originate from the same clone, they can inflate allele counts and bias genotype inference. If enabled, one representative sequence is selected per clone to represent that lineage in the analysis. This reduces the impact of clonal expansion, ensuring that allele frequencies reflect genuine germline variation rather than overrepresented clones, and minimizes noise from somatic hypermutation and sampling redundancy.

```{r clonal-representatives, eval=params$single_clone_representative}
if (params$single_clone_representative) {
  # Check that clone_id_column is provided
  if (is.null(params$clone_id_column) || params$clone_id_column == "") {
    stop("Error: 'clone_id_column' parameter must be provided when 'single_clone_representative' is TRUE.")
  }
  if (!params$clone_id_column %in% colnames(db)) {
    stop("'single_clone_representative' is TRUE but column '", params$clone_id_column,
         "' is not in the input. Run clonal assignment before genotyping, or set ",
         "'single_clone_representative' to FALSE.")
  }
  # Select clonal representatives
  db <- select_clonal_representatives(db,
    sequence_col = params$seq_column,
    germline_col = params$germline_column,
    clone_id_col = params$clone_id_column
  )
  db <- db %>% filter(clone_representative)
}
```   


### **This analysis is produced for subject `r paste(unique(db$subject_id), collapse=", ")` using the samples `r paste(unique(db$sample_id), collapse=", ")`.**

```{r genotype-method-heading, results='asis', echo=FALSE}
cat("# ", method_label, " genotype inference\n\n", sep = "")
cat(genotype_method_text(method, "description"), "\n")
```

```{r unmutated-note, eval=!find_unmutated, results='asis', echo=FALSE}
cat(
  "<div class=\"introduction_paragraph\">\n",
  "<strong>Recommendation:</strong> `find_unmutated` was not enabled, so the genotype below was inferred from all productive sequences, mutated ones included. ",
  "Inferring from unmutated sequences is generally preferable, because somatic hypermutation can make a sequence resemble an allele other than the one it rearranged from, which inflates support for alleles the subject may not carry.\n\n",
  "</div>\n",
  sep = ""
)
```

```{r genotype-inference}
# Method-specific arguments; only those of `method` are used.
priors <- list(
  v = as.numeric(strsplit(params$v_priors, ",")[[1]]),
  d = as.numeric(strsplit(params$d_priors, ",")[[1]]),
  j = as.numeric(strsplit(params$j_priors, ",")[[1]])
)

allele_thresholds <- read_allele_thresholds(params$allele_thresholds_db)

# allele_based: report threshold table coverage; stop if nothing matches.
threshold_coverage <- NULL
threshold_provenance <- NULL
if (identical(method, "allele_based")) {
  if (is.null(allele_thresholds)) {
    allele_thresholds_used <- .piglet_default_thresholds()
  } else {
    allele_thresholds_used <- allele_thresholds
  }
  threshold_provenance <- allele_threshold_provenance(
    thresholds = allele_thresholds_used,
    path = params$allele_thresholds_db,
    release = params$allele_thresholds_release
  )
  threshold_coverage <- allele_threshold_coverage(
    db = db,
    call_col = params[[paste0(segments_to_genotype[1], "_call")]],
    thresholds = allele_thresholds_used,
    asc_annotation = as.logical(params$asc_annotation),
    single_assignments = single_assignments
  )
  coverage_summary <- summarize_threshold_coverage(threshold_coverage)
  if (nrow(threshold_coverage) > 0 && coverage_summary$alleles_matched == 0) {
    stop(
      "None of the ", coverage_summary$alleles_observed, " alleles in the repertoire were ",
      "found in the allele threshold table, so every allele would fall back to ",
      "default_threshold = ", params$default_threshold, " and the population-derived ",
      "thresholds would be discarded. The table is matched on '",
      .threshold_key_column(allele_thresholds_used, as.logical(params$asc_annotation)),
      "'; check that asc_annotation matches the naming of the calls, or supply a ",
      "threshold table that carries the names used here."
    )
  }
  message("THRESHOLD_COVERAGE> ", coverage_summary$alleles_matched, "/",
          coverage_summary$alleles_observed, " alleles matched (",
          round(100 * coverage_summary$coverage), "%), ",
          coverage_summary$alleles_defaulted, " defaulted")
  message("THRESHOLD_SOURCE> ", threshold_provenance$source,
          " | release: ", threshold_provenance$release,
          " | md5: ", threshold_provenance$md5)
}

# Genotype each segment; output shares one schema across methods.
genotypes <- lapply(segments_to_genotype, function(seg) {
  infer_genotype(
    db = db,
    seg = seg,
    loci = loci,
    method = method,
    genotypeby = genotypeby,
    references = references,
    single_assignments = single_assignments,
    find_unmutated = find_unmutated,
    seq = params$seq_column,
    priors = priors[[seg]],
    fraction_to_explain = as.numeric(params$fraction_to_explain),
    gene_cutoff = as.numeric(params$gene_cutoff),
    allele_threshold_table = allele_thresholds,
    default_threshold = as.numeric(params$default_threshold),
    asc_annotation = as.logical(params$asc_annotation),
    translate_to_asc = as.logical(params$translate_to_asc),
    depth_adjusted_threshold = as.logical(params$depth_adjusted_threshold),
    z_score_threshold = as.numeric(params$z_score_threshold)
  )
})
genotypes <- do.call(rbind, genotypes)
if (is.null(genotypes) || nrow(genotypes) == 0) {
  stop(
    "Genotype inference with method '", method, "' produced no results for any of the ",
    "segments (", paste(segments_to_genotype, collapse = ", "), "). ",
    "See the warnings above for the underlying cause."
  )
}
genotypes <- augment_genotype_table(
  genotypes = genotypes,
  db_support = db_raw_productive,
  genotypeby = genotypeby,
  clone_id_column = params$clone_id_column,
  single_assignments = single_assignments,
  method = method
)
```

## Sequences in and out

<div class="introduction_paragraph">
Every filter applied before inference, with the rows it received and the rows it
passed on. A step that drops sequences is visible here; the clonal representative
selection below, when enabled, is counted the same way.
</div>

```{r filter-counts, results='asis', echo=FALSE}
tab_filters <- eetable(filter_steps,
  caption = "Sequences in and out of each filter applied before genotype inference",
  outdir = params$outdir,
  file = "tab_filters"
)
tab_filters$table
```

```{r echo=FALSE, results='asis'}
print_table_caption(tag = "tab_filters", text = tab_filters$caption)
```

```{r threshold-coverage-table, results='asis', echo=FALSE, eval=!is.null(threshold_coverage) && nrow(threshold_coverage) > 0}
tab_thresholds <- eetable(threshold_coverage,
  caption = paste0(
    "Allele threshold coverage: ", coverage_summary$alleles_matched, " of ",
    coverage_summary$alleles_observed, " alleles observed in the repertoire were found in the ",
    "threshold table (", round(100 * coverage_summary$coverage), "%). Alleles not found fall back to ",
    "default_threshold = ", params$default_threshold, ". Table in use: ",
    threshold_provenance$source, " (release ", threshold_provenance$release,
    ", md5 ", substr(threshold_provenance$md5, 1, 12), ")."
  ),
  outdir = params$outdir,
  file = "tab_threshold_coverage"
)
tab_thresholds$table
```

```{r echo=FALSE, results='asis', eval=!is.null(threshold_coverage) && nrow(threshold_coverage) > 0}
print_table_caption(tag = "tab_threshold_coverage", text = tab_thresholds$caption)
```

# Genotype inference results

## Genotype inference table
<div class="introduction_paragraph">
The genotype table summarizes the inferred alleles for each gene together with explicit support counts and the method's own confidence scores.
</div>

Interpret the columns as follows:

- `gene`: germline gene being genotyped.
- `genotyped_alleles`: allele set called into the genotype for that gene.
- `sequence_count`: number of productive sequences supporting each genotyped allele, comma-separated and in the same order as `genotyped_alleles`.
- `clone_count`: number of distinct clones among those same sequences, in the same order as `genotyped_alleles`. It is at most `sequence_count`, and equals it only where every supporting sequence belongs to a different clone. `NA` when the input carries no clone ID column.
- `note`: note or warning attached to the gene, when present.
- `candidate_alleles`: candidate allele set evaluated for that gene.
- `counts`: the allele counts the inference method itself computed, aligned with `candidate_alleles`. Both TIgGER and PIgLET *split* a sequence whose call names several alleles proportionally between those alleles, rather than giving each a full count. TIgGER rounds the result, so its values look like whole numbers; PIgLET does not, so its values are fractional.
- `total`: sum of `counts` across the gene's candidate alleles, on the same scale.

Columns specific to this method:

```{r genotype-columns-text, results='asis', echo=FALSE}
cat(genotype_method_text(method, "columns"), "\n")
```

```{r genotype-inference-table, results='asis'}
# Display results summary
caption <- paste0(method_label, " genotype inference results showing inferred genotypes")

tab_genotype <- eetable(genotypes,
  caption = caption,
  outdir = params$outdir,
  file = "tab_genotype"
)
tab_genotype$table
```

```{r echo=FALSE, results='asis'}
print_table_caption(tag="tab_genotype", text=tab_genotype$caption)
```

## Genotype inference plot

<div class="introduction_paragraph">
The genotype plot provides a visual summary of the inferred alleles for each gene. Each row corresponds to a gene, and the colored bars indicate which alleles are included in the inferred genotype. The plot is intended as a compact presence/absence view of the inferred genotype and should be interpreted together with the support counts in the table, and with the method's own confidence measure where it has one, rather than as a direct abundance plot.
</div>

```{r plot-genotype-inference, results='asis', echo=FALSE}
# Group plots by locus for mixed-locus data, otherwise by segment.
plot_groups <- split(
  genotypes,
  if (length(loci) > 1) substr(genotypes$gene, 1, 3) else substr(genotypes$gene, 4, 4)
)
genotype_plots <- lapply(plot_groups, plot_genotype, method = method)

emit_plot_sections(
  genotype_plots,
  prefix = "genotype-inference",
  caption_for = function(group) paste0(
    "Genotype inference summary: ", group,
    ". Each row represents a specific gene, where the colored bars indicate the inferred alleles."
  ),
  fig_height = pmax(vapply(plot_groups, nrow, integer(1)) * 0.5 + 1.5, 3)
)
```


```{r novel-allele-evidence, eval=has_novel, results='asis', echo=FALSE}
# load the novel allele evidence
load(params[["novel_allele_evidence"]])

# support both the current `novel` name and the legacy `novel_allele_db` name
evidence_table_exists <- exists("novel") || exists("novel_allele_db")
evidence_figures_exist <- exists("novel_allele_figures")

# check that both variables exist
if (evidence_table_exists && evidence_figures_exist) {
  if (exists("novel")) {
    novel_allele_db <- novel
  }

  # get the list of novel allele that passed the genotype
  novel_alleles_list <- genotypes %>%
    filter(!is.na(genotyped_alleles), genotyped_alleles != "") %>%
    select(gene, genotyped_alleles) %>%
    tidyr::separate_rows(genotyped_alleles, sep = ",") %>%
    mutate(
      genotyped_alleles = trimws(genotyped_alleles),
      polymorphism_call = paste0(gene, "*", genotyped_alleles)
    ) %>%
    filter(grepl("_", genotyped_alleles)) %>%
    pull(polymorphism_call) %>%
    unique()
  
  # filter the novel allele table
  novel_allele_db <- novel_allele_db %>%
    filter(polymorphism_call %in% novel_alleles_list)

  novel_allele_figures <- novel_allele_figures[names(novel_allele_figures) %in% novel_alleles_list]
  has_novel_table <- nrow(novel_allele_db) > 0
  has_novel_figures <- length(novel_allele_figures) > 0

  if (has_novel_table) {
    cat("## Novel allele evidence\n\n")
    cat("Candidate novel alleles shown below were first inferred upstream by TIgGER's `findNovelAlleles` procedure and are displayed here only if they were retained in the current genotype output.\n\n")
    cat("TIgGER identifies candidate novel alleles by examining each germline allele represented in the repertoire and testing whether specific positions show unexpectedly high mutation frequencies relative to the total mutation burden of the sequence. Positions with elevated baseline mutation frequencies are flagged as candidate polymorphic sites. The candidates shown in this report are therefore not every upstream candidate, but the subset that also passed into the genotype call.\n\n")
    cat("Additional filtering in the novel-allele report removes weak candidates, repeated-read SNP patterns, and duplicated polymorphism calls before evidence is passed to genotype. The table below lists the candidate novel alleles that remained consistent with the genotype result.\n\n")
    cat("### Novel allele table\n\n")
  } else {
    cat("## Novel allele evidence\n\n")
    cat("Candidate novel alleles were inferred upstream by TIgGER's `findNovelAlleles` procedure and compared against the genotype output here.\n\n")
    cat("A novel allele evidence file was provided, but no candidate novel alleles were retained in the current genotype call, so no novel allele table or evidence figures are shown.\n\n")
  }
  
} else {
  cat("## Novel allele evidence\n\n")
  cat("Warning: `novel`/`novel_allele_db` or `novel_allele_figures` was not found in `novel_allele_evidence`. Check the input file generated by the novel allele report.\n\n")
}

```

```{r novel-allele-table, eval=has_novel_table, results='asis', echo=FALSE}
caption <- "TIgGER candidate novel alleles retained in the genotype output"

tab_novel_allele <- eetable(novel_allele_db,
  caption = caption,
  outdir = params$outdir,
  file = "tab_novel_allele"
)
tab_novel_allele$table
```

```{r echo=FALSE, results='asis', eval=has_novel_table}
print_table_caption(tag="tab_novel_allele", text=tab_novel_allele$caption)
```

```{r novel-allele-figures-intro, eval=has_novel_figures, results='asis', echo=FALSE}
cat("### Novel allele figures\n\n")
cat("The figures below summarize the TIgGER evidence used to support each candidate novel allele. The first panel shows mutation frequency by position as a function of the total mutation count per sequence. In TIgGER's `plotNovel` output, red positions passed the novel-allele test, yellow positions passed the y-intercept screen but not all later criteria, and blue positions did not pass the y-intercept screen.\n\n")
cat("The second panel shows nucleotide usage at the candidate polymorphic positions across mutation-count bins. A consistent non-reference nucleotide among low-mutation sequences supports the interpretation that the pattern reflects a germline polymorphism rather than somatic mutation.\n\n")
cat("The third panel summarizes J gene and junction-length diversity among supporting sequences. Support coming from diverse rearrangements is less likely to reflect a single expanded clone, which helps distinguish true candidate novel alleles from clonal artefacts.\n\n")
```

```{r novel-allele-figures, eval=has_novel_figures, results='asis', echo=FALSE}
emit_plot_sections(
  novel_allele_figures,
  prefix = "novel-allele",
  caption_for = function(allele) paste0("Novel allele evidence: ", allele, "."),
  fig_height = 10
)
```

# Save genotype report

```{r save-genotypes}
# Save genotype report
genotypes_output_dir <- file.path(params$outdir, "genotypes")
dir.create(genotypes_output_dir, recursive = TRUE, showWarnings = FALSE)

write_genotypes(
  genotypes = genotypes,
  out_dir = genotypes_output_dir,
  out_fn = report_output_fn,
  genotypeby = genotypeby
)

# Save the thresholds used and their provenance.
if (!is.null(threshold_provenance)) {
  utils::write.table(threshold_provenance,
    file.path(genotypes_output_dir, sub("\\.tsv$", "_threshold_provenance.tsv", report_output_fn)),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
}
if (!is.null(threshold_coverage) && nrow(threshold_coverage) > 0) {
  utils::write.table(threshold_coverage,
    file.path(genotypes_output_dir, sub("\\.tsv$", "_threshold_coverage.tsv", report_output_fn)),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
}
```

# Build personal reference

We start from the full germline DB:
- For genes present in the genotype table, we remove all their alleles, then add back only the alleles validated by genotype (genotyped_alleles if set; otherwise the inferred alleles).
- For genes absent from the genotype table, we keep all alleles as in the original DB.

This yields a personal reference restricted to genotype-validated alleles for seen genes, while leaving unseen genes unchanged.

```{r build-reference, message=TRUE}
# create the germline output directory
references_dir <- file.path(imgt_db, species, "vdj")
# generate the personal genotyped reference
generate_genotyped_reference(
  genotypes = genotypes,
  references = references,
  output_dir = file.path(params$outdir, "references"),
  germline_dir = "db_genotype",
  references_dir = references_dir,
  loci = loci,
  species = species,
  genotypeby = genotypeby,
  quiet = FALSE
)
```

```{r log-end}
cat("START> Genotype Inference", file = log_fn, append = FALSE)
cat(paste0("\nSUBJECT> ", paste(unique(db$subject_id), collapse = ", ")), file = log_fn, append = TRUE)
cat(paste0("\nSAMPLE> ", paste(unique(db$sample_id), collapse = ", ")), file = log_fn, append = TRUE)
cat(paste0("\nMETHOD> ", method_label), file = log_fn, append = TRUE)
cat(paste0("\nFILE> ", basename(params$input)), file = log_fn, append = TRUE)
cat(paste0("\nGENOTYPE_LOCI> ", paste0(loci, collapse = ",")), file = log_fn, append = TRUE)
cat(paste0("\nGENOTYPE_SEGMENTS> ", paste0(segments_to_genotype, collapse = ",")), file = log_fn, append = TRUE)
cat(paste0("\nOUTPUT> ", basename(genotypes_output_dir)), file = log_fn, append = TRUE)
cat(paste0("\nOUTPUT_REFERENCE> ", basename(file.path(params$outdir, "references"))), file = log_fn, append = TRUE)
```

```{r, child=c('versions.Rmd')}
```
