---
title: "Immcantation - enchantR"
subtitle: "Reassign Alleles to Given Germline Alleles"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr reassign alleles to given germline alleles using TIgGER."
output: enchantr::immcantation
params:
   input:
      label: "`input`: Path to repertoires files"
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/refs/heads/airrflow/testdata-genotyping/airr_data_to_test_novel_alleles.tsv"
   imgt_db:
      label: "`imgt_db`: Path to IMGT reference  database, or a novel allele table (.tsv) to reassign only reads carrying a novel allele's parent."
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   species:
      label: "`species`"
      input: text
      value: "human"
   singlecell:
      label: "`singlecell`"
      input: text
      value: NULL
   outputby:
      label: "`outputby` name of the column in `input` that contains sample identifiers that will be used to split the output db." 
      input: text
      value: "sample_id"
   segments:
      label: "`segments`: which segments to reassign. Options are 'v', 'd', 'j'. Multiple segments can be provided as a comma-separated string. auto will reassign segments based on loci type."    
      input: text
      value: "auto"
   outname:
      label: "`outname`"
      input: text
      value: "allele-reassignment"
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: "`outdir`: Output directory"
      input: text
      value: !r file.path(getwd(),'enchantr')
   date:
      label: "`date`: Run date"
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo:
      label: "`echo`: Show code in the report."
      input: checkbox
      value: TRUE
   cache:
      label: "`cache`: Use cached results"
      input: checkbox
      value: FALSE
editor_options:
   chunk_output_type: console
---

<div class="introduction_paragraph">
This report performs novel allele inference using TIgGER to identify potential new immunoglobulin alleles in repertoire sequencing data. Novel alleles are detected by identifying positions with unexpectedly high mutation frequencies compared to the germline reference database.
</div>

```{r global-options, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4, fig.path = "figures/",
                      echo = params$echo, cache = params$cache,
                      warning = FALSE, message = FALSE,
                      eval.after = "fig.cap",
                      out_dir = params$outdir,
                      eval.opts = c("eval", "echo", "fig.height", "fig.width"))

# Libraries
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("tigger"))
suppressPackageStartupMessages(library("alakazam"))

if (!dir.exists(params$outdir)) {
  dir.create(params$outdir, recursive = TRUE)
}

file.copy(params$logo,
          file.path(params$outdir, "assets", "logo.png"),
          recursive = TRUE, overwrite = TRUE)

```


# Input parameters

```{r input-parameters, results = 'asis'}
printParams(params)
```
```{r input-parameters-save, echo=FALSE, results = 'asis'}
print_table_caption(tag = "input", text = "Input parameters.")
save(params, file = file.path(params$outdir, "params.RData"))
```

# Read repertoires
<div class="introduction_paragraph">
Tables summarizing number of sequences in the provided repertoires according to different characteristics.
</div>

```{r}
# Read repertoire
db <- readInput(params[['input']], col_select = NULL)
singlecell <- params$singlecell

if (params$species == "auto") {
  if ("species" %in% colnames(db)) {
    species <- unique(db[["species"]])
    if (length(species) > 1) {
      stop("Multiple species detected. Only one allowed.")
    }
  } else {
    stop("Can't detect species. The column `species` does not exist in `db`.")
  }
} else {
  species <- params$species
}

# Check for single cell label
if (!is.null(singlecell)) {
  if (singlecell %in% colnames(db)) {
    db[[ singlecell ]] <- as.logical(db[[singlecell]])
  } else {
    stop("`",singlecell, "` is not a valid field in `db`.")
  }
} else {
  singlecell <- "single_cell"
  db[[singlecell]] <- FALSE
  if ("cell_id" %in% colnames(db)) {
    message("Setting `singlecell` using `cell_id`.")
    db[[singlecell]][!db[["cell_id"]] %in% c(NA, "")] <- TRUE
  }

}

if (!"locus" %in% colnames(db)) {
  db[["locus"]] <- getLocus(db[["v_call"]])
}
heavy_chains <- isHeavyChain(db[["locus"]])

if (params$segments == "auto") {
  segments_to_reassign <- if (any(heavy_chains)) c("v", "d", "j") else c("v", "j")
} else {
  segments_to_reassign <- unlist(strsplit(params$segments, ","))
  segments_to_reassign <- tolower(trimws(segments_to_reassign))
  valid_segments <- c("v", "d", "j")
  if (!all(segments_to_reassign %in% valid_segments)) {
    stop("Invalid segments provided in `segments` parameter. Valid options are 'v', 'd', 'j'.")
  }
  if (any(segments_to_reassign == "d") && !any(heavy_chains)) {
    stop("D segments can only be reassigned for heavy chain repertoires.")
  }
}

subject_levels <- str_sort(unique(db$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(db$sample_id),  numeric = TRUE)
db <- db %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id)
```


```{r}
input_size <- nrow(db)
input_sizes <- db %>%
  group_by(!!!rlang::syms(unique(c("input_file")))) %>%
  summarize(input_size = n())

# Track input size of the expected output groups, for the
# end of report summary
input_sizes_byoutput <- db %>%
  group_by(!!!rlang::syms(unique(params$outputby))) %>%
  mutate(
    num_input_files = length(unique(input_file)),
    input_files = paste(unique(input_file), collapse = ",")
  ) %>%
  ungroup() %>%
  group_by(!!!rlang::syms(unique(c(params$outputby, "input_file", "num_input_files", "input_files")))) %>%
  summarize(
    input_size = n()
  )

```

# Read Germline database

Import the nucleotide germline reference used for alignment.
Using the same reference ensures positional numbering and mutation counts remain consistent with the upstream alignment step. Each `v_call` present in the repertoire should have a corresponding entry in the germline database.

```{r read-germline}
loci <- unique(db[["locus"]])
# a novel allele table instead of a germline database
novel_table <- grepl("\\.tsv$", params$imgt_db)
if (novel_table) {
  novel <- read.delim(params$imgt_db, colClasses = "character")
  novel_cols <- c("germline_call", "germline_imgt", "polymorphism_call", "novel_imgt")
  if (!all(novel_cols %in% colnames(novel))) {
    stop("`imgt_db` is a .tsv file but not a novel allele table. Missing columns: ",
         paste(setdiff(novel_cols, colnames(novel)), collapse = ", "))
  }
} else {
  # download and unzip if needed
  imgt_db <- prepareIMGT(params$imgt_db)
  references <- dowser::readIMGT(file.path(imgt_db, species, "vdj"), quiet = TRUE)
}
```

# Reassign sequences to germline database

Sequences are re-aligned to the updated germline database to include the novel alleles in the repertoire annotations.

For re-alignment the `reassignAlleles` function from TIgGER is used. The function utilizes hamming distance on the gapped sequence (i.e `sequence_alignment`), 

in order to save time, initial gene assignments are preserved and the allele calls are chosen from among those provided in `germline_db`, based on a simple alignment to the sample sequence.

When a novel allele table is provided instead of a germline database, only sequences assigned to the allele a novel allele was inferred from are reassigned, choosing between that allele and its novel alleles. All other assignments are kept.

```{r reassign, message = TRUE, warning = TRUE}
if (novel_table) {
  # novel alleles and the alleles they were inferred from
  novel_db <- c(setNames(novel[["germline_imgt"]], novel[["germline_call"]]),
                setNames(gsub("-", ".", novel[["novel_imgt"]], fixed = TRUE), novel[["polymorphism_call"]]))
  novel_db <- novel_db[!duplicated(names(novel_db))]
  # only sequences assigned to a parent of a novel allele are reassigned
  has_parent <- vapply(strsplit(db[["v_call"]], ","),
                       function(x) any(x %in% novel[["germline_call"]]), logical(1))
  message("Reassigning ", sum(has_parent), " sequences assigned to a parent of a novel allele.")
  if (any(has_parent)) {
    # compare over the V alignment when its germline coordinates are available
    trim_seq <- all(c("v_germline_start", "v_germline_end") %in% colnames(db))
    db[has_parent, ] <- tigger::reassignAlleles(db[has_parent, ], novel_db,
                                                trim_seq = trim_seq, overwrite = TRUE)
  }
} else if (!all(is.na(references))) {
  # Reassign per locus so each repertoire is only compared against its own
  # locus' alleles, then recombine in the original row order.
  db <- reassign_loci(db, loci, segments_to_reassign, references)
}
```

# Final repertoires

Find your final repertoires here:

```{r fn, results='asis', echo=F}

if (!is.null(params$outname)) {
  output_fn <- paste0(params$outname, "_reassign-pass.tsv.gz")
} else {
  output_fn <- sub("\\.tsv(\\.gz)?$", "_reassign-pass.tsv.gz", basename(params$input))
}

if (!is.null(params$log)) {
  log_fn <- file.path(params$outdir, paste0(params$log, ".txt"))
} else {
  log_fn <- file.path(params$outdir, sub("_reassign-pass.tsv.gz$", "_command_log.txt", basename(output_fn)))
}
```

```{r dbpass, results='asis', echo=FALSE}

# collpase in case there are multiple rows
# due to multiple input files
input_sizes_byoutput <- input_sizes_byoutput %>%
  select(-input_files) %>%
  ungroup() %>%
  group_by(!!!rlang::syms(params$outputby)) %>%
  summarise(across(everything(), function(x) {
    paste(x, collapse = ",")
  }))

output_groups <- db %>%
  ungroup() %>%
  select(!!!rlang::syms(c(params$outputby, "input_file"))) %>%
  group_by(!!!rlang::syms(params$outputby)) %>%
  summarize(output_size = n()) %>%
  right_join(input_sizes_byoutput) %>%
  select(input_size, output_size, everything())

output_files <- c()
output_sizes <- c()
input_sizes <- c()
dir.create(file.path(params$outdir, "repertoires"))
for (i in seq_len(nrow(output_groups))) {
  group_db <- left_join(output_groups[i, params$outputby, drop = FALSE],
                        db,
                        by = c(params$outputby))

  group_db_label <- gsub("^id_", "", makeLabel(group_db, fields = params$outputby))
  group_fn <- file.path(params$outdir, "repertoires", paste0(group_db_label, "_", output_fn))
  output_files <- c(output_files, group_fn)
  output_sizes <- c(output_sizes, nrow(group_db))
  i_input_size <- sum(as.numeric(strsplit(output_groups[["input_size"]][i], ",")[[1]]))
  input_sizes <- c(input_sizes, i_input_size)
  write_rearrangement(group_db %>% select(-input_file), file = group_fn)
  cat("- [", group_db_label, "](", file.path("repertoires", basename(group_fn)), ")\n", sep = "")
}
```

```{r log, echo=FALSE}
cat("START> ReassignAlleles", file = log_fn, append = FALSE)
cat(paste0("\nFILE-", i, "> ", basename(output_groups[["input_file"]][i])), file = log_fn, append = TRUE)
cat(paste0("\nOUTPUT-", i, "> ", basename(output_files[i])), file = log_fn, append = TRUE)
```

```{r, child=c('versions.Rmd')}
```