--- 
title: "Immcantation - enchantR"
subtitle: "Find Threshold"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr find threshold report"
output: enchantr::immcantation
params:
   input: 
      label: "`input`: Path to repertoires file"
      input: file
      value: "https://yale.box.com/shared/static/dajikxtvbc3c5yw1o2v69fjtx5bh58r2.tsv"
   cloneby:
      label: "`cloneby`"
      input: text
      value: "sample_id"   
   crossby:
      label: "`crossby`"
      input: text
      value: "sample_id"
   findthreshold_method:
      label: "`findthreshold_method`"
      input: text
      value: "gmm"
   findthreshold_model:
      label: "`findthreshold_model`"
      input: text
      value: "gamma-norm"
   findthreshold_edge:
      label: "`findthreshold_edge`"
      input: numeric
      value: 0.9
   findthreshold_cutoff:
      label: "`findthreshold_cutoff`"
      input: text
      value: "user"
   findthreshold_spc:
      label: "`findthreshold_spc`"
      input: numeric
      value: 0.995
   singlecell:
      label: "`singlecell`"
      input: text
      value: NULL          
   subsample:
      label: "`subsample`"
      input: numeric
      value: 0
   outname:
      label: "`outname`"
      input: text
      value: "find-threshold"
   nproc:
      label: "`nproc`"
      input: numeric
      value: 1      
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: '`outdir`: Output directory'
      input: text
      value: !r file.path(getwd(),'enchantr')
   date: 
      label: '`date`: Run date'
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo: 
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options: 
  chunk_output_type: console
---

```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4, fig.path = "figures/",
                      echo=params$echo,cache=params$cache,
                      warning=FALSE, message=FALSE,
                      eval.after="fig.cap",
                      out_dir=params$outdir,
                      eval.opts = c('eval', 'echo', 'fig.height', 'fig.width'))
options(rlang_backtrace_on_error_report = "full")


# Libraries
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("alakazam"))
suppressPackageStartupMessages(library("shazam"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("plotly"))
suppressPackageStartupMessages(library("R.utils"))

if (!dir.exists(params[['outdir']])) {
    dir.create(params[['outdir']], recursive = T)
}

file.copy(params$logo,
          file.path(params$outdir,"assets", "logo.png"),
          recursive = T, overwrite=T)
```

# Input 

## Parameters


```{r input-parameters, results='asis'}
printParams(params)
```
```{r echo=FALSE, results='asis'}
print_table_caption(tag="input", text="Input parameters.")
save(params, file=file.path(params$outdir,"params.RData"))
```

## Data Sources

```{r}
# Read repertoires, 
# keep only needed columns
selected_columns <- unique(c("sample_id", "sequence_id", 
                      params$cloneby,
                      params$crossby,
                      "v_call", "j_call", "junction", 
                      "cell_id", params$singlecell, "locus"))
db <- readInput(params[['input']], col_select=selected_columns)
input_sizes <- db %>%
    count(input_file)
input_sizes <- input_sizes %>% rename("sequences" = "n")
eetable(input_sizes)$table
```

```{r}
input_size <- nrow(db)
is_heavy <- isHeavyChain(db[['locus']])
singlecell <- params$singlecell

# Check for single cell label
if (!is.null(singlecell)) {
    if (singlecell %in% colnames(db)) {
         db[[ singlecell ]] <- as.logical(db[[ singlecell ]] )
    } else {
        stop("`", singlecell, "` is not a valid field in `db`.")
    }
} else {
   singlecell <- "single_cell"
   db[[singlecell]] <- F
   if ("cell_id" %in% colnames(db)) {
       message("Setting `singlecell` using `cell_id`.") 
       db[[singlecell]][!db[['cell_id']] %in% c(NA, '')] <- T
   }

}

na_single <- is.na(db[[singlecell]])
if (sum(na_single)>0) {
    warning(sum(na_single), " sequences are missing single cell information. Using single_cell=F")
    db[[singlecell]][na_single] <- FALSE
}


bulk_heavy <- is_heavy & !db[[singlecell]]
sc_heavy <- is_heavy & db[[singlecell]]
bulk_dtn <- data.frame()
sc_dtn <- data.frame()
subsample <- params[['subsample']]
if (subsample %in% c(NA, 0, NULL)) {
    subsample <- NULL
}
```

Number of sequences loaded: `r input_size`.

```{r, results='asis'}
input_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "locus")))) %>%
   summarize(n=n(), .groups="drop") %>%
       pivot_wider(names_from=locus, values_from=n)

input_summary_table <- eetable(input_summary) 
input_summary_table$table
```
```{r echo=FALSE,results='asis'}
print_table_caption(tag="input_summary", text="Input data summary.")
```

# Calculate Threshold for Clonal Relationships

The `shazam` package provides methods to find an appropriate distance threshold to determine clonal relationships for each dataset (`distToNearest` and `findThreshold`) by generating a distribution of distances between each sequence and its closest non-identical neighbor. Typically, the distance-to-nearest distribution for a repertoire is bimodal. The first mode (on the left) represents sequences that have at least one clonal relative in the dataset, while the second mode (on the right) is representative of the sequences that do not have any clonal relatives in the data. A reasonable threshold will separate these two modes of the distribution. The distance may be selected manually, or `findThreshold` can be used to automatically select a threshold.

Specifying the `cross` argument to `distToNearest` forces distance calculations to be performed across groups, such that the nearest neighbor of each sequence will always be a sequence in a different group. 

```{r sum_heavy_0, eval=sum(is_heavy)==0, echo=sum(is_heavy)==0, results='asis'}
cat("0 heavy chain sequences present. `distToNearest` requires heavy chain data. Setting `dist_neartest` to 'NA'")
db$dist_nearest <- NA
```

```{r distToNearest_bulk, eval=(any(db[[singlecell]] == F) && sum(bulk_heavy)>0) }
if (any(db[[singlecell]] == F) && sum(bulk_heavy)>0) {
    message("Using ",sum(bulk_heavy), " bulk heavy chain sequences.")

    # Add distances to nearest neighbor within a "cloneby" group
    bulk_dtn <- distToNearest(db[bulk_heavy,] , 
                               fields=params$cloneby,
                               sequenceColumn="junction", 
                               vCallColumn="v_call", jCallColumn="j_call",
                               model="ham", first=FALSE, VJthenLen=TRUE, normalize="len",
                               nproc=params$nproc,
                               subsample = subsample)
    
    bulk_dtn <- distToNearest(bulk_dtn, 
                               fields=NULL,
                               cross=params$crossby,
                               sequenceColumn="junction", 
                               vCallColumn="v_call", jCallColumn="j_call",
                               model="ham", first=FALSE, VJthenLen=TRUE, normalize="len",
                               nproc=params$nproc,
                               subsample = subsample)
        
} 
```

```{r distToNearest_single, eval=(any(db[[singlecell]] == T) && sum(sc_heavy)>0) }
if (any(db[[singlecell]] == T) && sum(sc_heavy)>0) {
    message("Using ",sum(sc_heavy), " sc heavy chain sequences.")

    # Create unique cell_id
    # db[['cell_id']] <- paste0(db[['sample_id']],db[['cell_id']])

    # Add distances to nearest neighbor within a "cloneby" group
    sc_dtn <- distToNearest(db[sc_heavy,] , 
                          fields=params$cloneby,
                          sequenceColumn="junction", 
                          vCallColumn="v_call", jCallColumn="j_call",
                          model="ham", first=FALSE, VJthenLen=TRUE, normalize="len",
                          nproc=params$nproc,
                          cellIdColumn = "cell_id",
                          locusColumn = "locus",
                          onlyHeavy = TRUE,
                          subsample = subsample)
    # Add distances to nearest neighbor across a "crossby" group
    sc_dtn <- distToNearest(sc_dtn, 
                          fields=NULL,
                          cross=params$crossby,
                          sequenceColumn="junction", 
                          vCallColumn="v_call", jCallColumn="j_call",
                          model="ham", first=FALSE, VJthenLen=TRUE, normalize="len",
                          nproc=params$nproc,
                          cellIdColumn = "cell_id",
                          locusColumn = "locus",
                          onlyHeavy = TRUE,
                          subsample = subsample)    
} 
```

## `distToNearest` plot

```{r dtnplot, fig.width=8, fig.height=max(4, nrow(unique(db[,c("sample_id", singlecell)], drop=F))/2), fig.cap=caption}

caption <- paste0("Distribution of the distance to the nearest sequence within ",
            paste(params$cloneby, collapse=","),", displayed by sample_id. When ",
            "cross ",paste(params$crossby, collapse=","),
            " distances are available, they are shown in an inverse y-axis.")
if (!is.null(subsample)) {
    caption <- paste0(caption," Subsampling requested: ", subsample, ".")
}

wrap_formula <- as.formula(paste0(
                        c(
                            singlecell,
                            "~",
                            paste(unique(c("sample_id", params$clone_by)), collapse = "+")
                        ),collapse="")
)
db <- bind_rows(bulk_dtn,sc_dtn)

dtnplot <- NULL

if (any(!is.na(db[['dist_nearest']]))) {
    dtnplot <- ggplot(db) +
        geom_histogram(aes(dist_nearest, fill=sample_id),binwidth = 0.01) +
        scale_x_continuous(
            breaks = seq(0, 1, 0.1)
        )
} else {
    cat("All `dist_nearest` values are NA.")
    warning("All `dist_nearest` values are NA.")
}

if ("cross_dist_nearest" %in% c(colnames(bulk_dtn),colnames(sc_dtn))) {
   if (any(!is.na(db[['cross_dist_nearest']]))) {
    dtnplot <- dtnplot + 
        geom_histogram(
            aes(x=cross_dist_nearest, y=-(..count..), fill=sample_id), 
            binwidth=0.01, position="identity") +
            #scale_y_continuous(labels = abs) +
            ylab("count")
   } else {
      cat("All `cross_dist_nearest` values are NA.")
      warning("All `cross_dist_nearest` values are NA.")
   }
}

if (is.null(dtnplot)) {
    cat("All distance values are NA.")
}
```

```{r showdtnplot, eval=!is.null(dtnplot), fig.width=8, fig.height=min(50, max(4, nrow(unique(db[,c("sample_id", singlecell)], drop=F))/1.5)), fig.cap=caption}
dtnplot <- dtnplot +
    facet_wrap(wrap_formula, scales = "free_y", ncol=2, labeller = "label_both") +
    expand_limits(x = 0, y =0) +
    theme_enchantr() +
    theme(legend.position = "bottom")

dtnplot <- eeplot(dtnplot, 
                  outdir=params$outdir,
                  file=knitr::opts_current$get('dtnplot'),
                  caption=caption
)
ggplotly(dtnplot + theme(panel.spacing=unit(2, 'lines'), legend.position="none"))
caption <- dtnplot$enchantr$html_caption
```

## Find threshold

`findThreshold` uses the distribution of distances calculated in the previous step to determine an appropriate threshold for the dataset. 


### Threshold(s) summary table

```{r, eval=all(is.na(db[['dist_nearest']]))}
# This chunk runs when all dist_nearest values are NA
threshold <- NA
threshold_summary <- data.frame(mean_threshold = NA)
cat("All `dist_nearest` values are NA. Skipping threshold analysis.")
```

```{r threshold, eval=!all(is.na(db[['dist_nearest']])), results='asis'}
cross_dist_nearest <- NULL
if ("cross_dist_nearest" %in% colnames(db)) {
    if (any(!is.na(db[['cross_dist_nearest']]))) {
       cross_dist_nearest <- "cross_dist_nearest"
    } else {
       warning("All `cross_dist_nearest` values are NA.")
    }
}
threshold <- findThresholdDb(db, 
                            distanceColumn="dist_nearest", 
                            crossDistanceColumn=cross_dist_nearest,
                            method=params$findthreshold_method, 
                            model=params$findthreshold_model,
                            edge=params$findthreshold_edge, 
                            cutoff = params$findthreshold_cutoff, 
                            spc = params$findthreshold_spc,
                            nproc=params$nproc, 
                            fields=params$cloneby,
                            subsample=subsample)

threshold_summary <- gmmSummary(threshold) %>%
    mutate(mean_threshold=mean(threshold, na.rm=TRUE))
mean_threshold <- round(threshold_summary$mean_threshold[1],2)

tab_caption <- paste0("Summary of threshold values, p-value from Hartigans’ dip statistic (HDS) test with <0.05 indicating significant bimodality. Mean threshold is: ", mean_threshold)
if (!is.null(subsample)) {
    tab_caption <- paste0(tab_caption," Subsampling requested: ", subsample, ".")
}

tab <- eetable(threshold_summary, 
               outdir=params$outdir, 
               file=paste(params$outname, "threshold-summary", sep="_"),
               caption=tab_caption)  
tab$table %>%
        DT::formatRound(columns=setdiff(colnames(threshold_summary)[-1],"model"),
                    digits=3)
```
```{r echo=FALSE,eval=!all(is.na(db[['dist_nearest']])),results='asis'}
print_table_caption(tag="threshold_summary",text=tab$caption)
```

```{r, eval=!all(is.na(db[['dist_nearest']])), results='asis', fig.width=3.5, fig.height=3, fig.cap=caption_list}
caption_list <- NULL
tmp <- lapply(threshold, function(thr) {
    this_thr <- NA
    if (!is.null(attributes(thr$GmmThreshold))) {
        this_thr <- round(attributes(thr$GmmThreshold)$threshold,2)
    }
    cat ("\n\n### ",paste(c(params$crossby, " ", thr[['fields']], ", estimated threshold:", this_thr),collapse=" "),"\n\n")
    fn <- paste(c(thr[['fields']],"dtnthrplot"),collapse="_")
    dtnthrcaption <- paste0("Distribution of the distance to the nearest sequence within ",
            paste(params$cloneby, collapse=",")," (",
            paste(thr[['fields']],collapse="_")
            ,"). When ",
            "cross ",paste(params$crossby, collapse=","),
            " distances are available, they are shown in an inverse y-axis.")
    p <- thr[['plot']] +
        scale_x_continuous(
            breaks = seq(0, 1, 0.1)
        )
    p <- eeplot(p, 
                  outdir=params$outdir, 
                  file=fn,
                  caption=dtnthrcaption
    )
    print(p)
    caption_list <<- c(caption_list, p$enchantr$html_caption)    
})
```

# Save

```{r dbpass}
# pass <- db[['collapse_pass']]

if (!is.null(params$outname)) {
    output_fn <- paste(params$outname,"threshold-pass.tsv", sep="_")
} else {
    output_fn <- sub(".tsv$", "_threshold-pass.tsv", basename(params$input))
}

if (!is.null(params$log)) {
    log_fn <- paste0(params$log,".txt")
} else {
    log_fn <- sub("threshold-pass.tsv$", "command_log.txt", basename(output_fn))
}

# write_rearrangement(db , file=output_fn)
cat(threshold_summary[['mean_threshold']][1], 
    file=file.path(params$outdir,"tables",sub("threshold-pass.tsv$", "threshold-mean.tsv" ,output_fn)),
    append=F
)
```

```{r log}
cat("START> FindThreshold", file=log_fn, append=F)
cat(paste0("\nFILE> ",basename(params$input)), file=log_fn, append=T)
cat(paste0("\nOUTPUT> ",basename(output_fn)), file=log_fn, append=T)
cat(paste0("\nPASS> ",nrow(db)), file=log_fn, append=T)
cat(paste0("\nFAIL> ",input_size-nrow(db)), file=log_fn, append=T)
```


```{r, child=c('versions.Rmd')}
```
