---
title: "Immcantation - enchantR"
subtitle: "Novel allele inference"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr novel allele inference"
output: enchantr::immcantation
params:
   input:
      label: "`input`: Path to repertoires files"
      input: file
      value: "https://yale.box.com/shared/static/dajikxtvbc3c5yw1o2v69fjtx5bh58r2.tsv"
   imgt_db:
      label: "`imgt_db`: Path to IMGT reference  database."
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   species:
      label: "`species`"
      input: text
      value: "human"
   germline_min:
      label: "`germline_min`"
      input: numeric
      value: 200
   min_seqs:
      label: "`min_seqs`"
      input: numeric
      value: 50
   auto_mutrange:
      label: "`auto_mutrange`"
      input: checkbox
      value: TRUE
   mut_range:
      label: "`mut_range`"
      input: text
      value: "1:10"
   pos_range:
      label: "`pos_range`"
      input: text
      value: "1:318"
   pos_range_max:
      label: "`pos_range_max`"
      input: numeric
      value: NULL
   y_intercept:
      label: "`y_intercept`"
      input: numeric
      value: 0.125
   alpha:
      label: "`alpha`"
      input: numeric
      value: 0.05
   j_max:
      label: "`j_max`"
      input: numeric
      value: 0.15
   min_frac:
      label: "`min_frac`"
      input: numeric
      value: 0.75
   v_call:
      label: "`v_call`: Column name for V gene calls"
      input: text
      value: "v_call"
   j_call:
      label: "`j_call`: Column name for J gene calls"
      input: text
      value: "j_call"
   seq:
      label: "`seq`: Column name for sequence alignment"
      input: text
      value: "sequence_alignment"
   junction:
      label: "`junction`: Column name for junction sequence"
      input: text
      value: "junction"
   junction_length:
      label: "`junction_length`: Column name for junction length"
      input: text
      value: "junction_length"
   reassign:
      label: "`reassign`: Reassign sequences to personal genotype"
      input: checkbox
      value: FALSE
   outname:
      label: "`outname`"
      input: text
      value: "tigger"
   nproc:
      label: "`nproc`"
      input: numeric
      value: 1
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: "`outdir`: Output directory"
      input: text
      value: !r file.path(getwd(),'enchantr')
   date:
      label: "`date`: Run date"
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo:
      label: "`echo`: Show code in the report."
      input: checkbox
      value: TRUE
   cache:
      label: "`cache`: Use cached results"
      input: checkbox
      value: FALSE
editor_options:
   chunk_output_type: console
---

<div class="introduction_paragraph">
This report performs novel allele inference using TIgGER to identify potential new immunoglobulin alleles in repertoire sequencing data. Novel alleles are detected by identifying positions with unexpectedly high mutation frequencies compared to the germline reference database.
</div>

```{r global-options, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4, fig.path = "figures/",
                      echo = params$echo, cache = params$cache,
                      warning = FALSE, message = FALSE,
                      eval.after = "fig.cap",
                      out_dir = params$outdir,
                      eval.opts = c("eval", "echo", "fig.height", "fig.width"))

# Libraries
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("tigger"))
suppressPackageStartupMessages(library("alakazam"))

if (!dir.exists(params$outdir)) {
  dir.create(params$outdir, recursive = TRUE)
}

file.copy(params$logo,
          file.path(params$outdir, "assets", "logo.png"),
          recursive = TRUE, overwrite = TRUE)

# make sure mut_range is range
mut_range <- params$mut_range
if (is.character(mut_range)) {
   if (grepl(":", mut_range)) {
      mut_range <- unlist(strsplit(mut_range, ":"))
      mut_range <- mut_range[1]:mut_range[2]
   } else {
      warning("Couldn't parse mut_rage; reverting to default 1:10")
      mut_range <- 1:10
   }
}

# make sure pos_range is range
pos_range <- params$pos_range
if (is.character(pos_range)) {
   if (grepl(":", pos_range)) {
      pos_range <- unlist(strsplit(pos_range, ":"))
      pos_range <- pos_range[1]:pos_range[2]
   } else {
      warning("Couldn't parse pos_range; reverting to default 1:318")
      pos_range <- 1:318
   }
}

```


```{r helper-functions, include = FALSE, cache = FALSE}

# Helper for light chain inference by gene range
infer_novel_light_chain <- function(db, germline_db, params) {
  gene_range <- tigger_upper_bound(db)
  purrr::map_dfr(seq_along(gene_range), function(i) {
    upper_range <- as.numeric(names(gene_range)[i])
    genes <- gene_range[i]
    sub_db <- db[stringi::stri_detect_regex(db[[params$v_call]], genes), ]
    if (nrow(sub_db) == 0) return(NULL)
    low_range <- min(sub_db$v_germline_start)
    novel_df <- try(findNovelAlleles(
      data = sub_db,
      germline_db = germline_db,
      pos_range = low_range:upper_range,
      v_call = params$v_call,
      j_call = params$j_call,
      seq = params$seq,
      junction = params$junction,
      junction_length = params$junction_length,
      nproc = params$nproc,
      germline_min = params$germline_min,
      min_seqs = params$min_seqs,
      y_intercept = params$y_intercept,
      alpha = params$alpha,
      j_max = params$j_max,
      min_frac = params$min_frac,
      auto_mutrange = params$auto_mutrange,
      mut_range = mut_range
    ))
    if (inherits(novel_df, "try-error")) return(NULL)
    novel_df
  })
}


```


# Input parameters

```{r input-parameters, results = 'asis'}
printParams(params)
```
```{r input-parameters-save, echo=FALSE, results = 'asis'}
print_table_caption(tag = "input", text = "Input parameters.")
save(params, file = file.path(params$outdir, "params.RData"))
```

# Read repertoires

Load repertoire annotation tables in AIRR format.
Expected columns include `sequence_id`, `productive`, `v_call`, `j_call`, `sequence_alignment`, `junction`, and `junction_length`. Only productive, in-frame sequences without stop codons are retained to avoid artefacts.

```{r read-input}
# Read repertoire

cols_to_read <- unique(c("sequence_id", "productive",
                         params$seq, params$v_call, params$j_call,
                         params$junction, params$junction_length))

db <- read_rearrangement(params[["input"]], col_select = cols_to_read)

if (!"locus" %in% colnames(db)) {
  db[["locus"]] <- getLocus(db[[params$v_call]])
}

if (!is.null(params$outname)) {
  output_fn <- paste0(basename(params$outname), "_novel-pass.tsv.gz")
  report_output_fn <- paste0(basename(params$outname), "_novel_report.tsv")
} else {
  output_fn <- sub(".tsv$", "_novel-pass.tsv", basename(params$input))
  report_output_fn <- sub(".tsv$", "_novel_report.tsv", basename(params$input))
}

if (!is.null(params$log)) {
  log_fn <- file.path(params$outdir, paste0(params$log, ".txt"))
} else {
  log_fn <- file.path(params$outdir, sub("_novel-pass.tsv$", "_command_log.txt", basename(output_fn)))
}

db <- db %>% dplyr::filter(.data[["productive"]] == TRUE)
```

# Read Germline database

Import the nucleotide germline reference used for alignment.
Using the same reference ensures positional numbering and mutation counts remain consistent with the upstream alignment step. Each `v_call` present in the repertoire should have a corresponding entry in the germline database.

```{r read-germline}

loci <- unique(db[["locus"]])
call_tag <- paste0(loci, toupper(substr(params$v_call, 1, 1)))
ref_output_fn <- paste0(params$species, "_", call_tag, "_novel.fasta")


# If it is a zip file or URL, run prepareIMGT; else use as is
if (grepl("\\.zip$", params$imgt_db) || grepl("^https?://", params$imgt_db)) {
  imgt_db <- prepareIMGT(params$imgt_db)
  germline_db <- dowser::readIMGT(file.path(imgt_db, params$species, "vdj"), quiet = TRUE)
   ## filter the germline database to include only alleles of params$v_call
   germline_db <- unlist(lapply(loci, function(locus) germline_db[[locus]][[toupper(substr(params$v_call, 1, 1))]]))
} else {
  germline_db <- tigger::readIgFasta(params$imgt_db)
  germline_db <- germline_db[grepl(toupper(substr(params$v_call, 1, 1)), names(germline_db))]
}

```

```{r log}
cat("START> InferNovelAlleles", file = log_fn, append = FALSE)
cat(paste0("\nFILE> ", basename(params$input)), file = log_fn, append = TRUE)
```

# Run `findNovelAlleles`

`findNovelAlleles` evaluates each germline allele represented in the dataset to detect potential polymorphic positions.
Mutation frequency at every IMGT-numbered site is modeled as a function of the overall mutation load per sequence.
Positions with a disproportionately high baseline mutation frequency (high y-intercept) are flagged as possible germline differences.

*Key parameters*:

- `germline_min`: excludes alleles with too few sequences for statistical testing.

- `min_seqs`: ensures adequate read depth in the target mutation and positional ranges.

- `auto_mutrange` / `mut_range`: controls which sequences (by mutation count) are included to balance sensitivity vs noise.

- `pos_range`: limits the region analyzed to reliably aligned V-segment positions, avoiding false SNPs near the CDR3 junction.

*Expected*: only a small fraction of alleles show candidate polymorphic sites; most known alleles return no novel calls.

```{r infer-novel-alleles}

# Run TIgGER novel allele inference
if (all(loci == "IGH")) {
  # Heavy chain: run as usual
  novel <- try(findNovelAlleles(
    data = db,
    germline_db = germline_db,
    v_call = params$v_call,
    j_call = params$j_call,
    seq = params$seq,
    junction = params$junction,
    junction_length = params$junction_length,
    germline_min = params$germline_min,
    min_seqs = params$min_seqs,
    y_intercept = params$y_intercept,
    alpha = params$alpha,
    j_max = params$j_max,
    min_frac = params$min_frac,
    auto_mutrange = params$auto_mutrange,
    mut_range = mut_range,
    pos_range = pos_range,
    nproc = params$nproc
  ))

} else {
  # Light chain: run per gene-range
  novel <- infer_novel_light_chain(db, germline_db, params)
}
```

# Filter and clean novel alleles

*Per-gene cap (max two alleles per V gene)*

We retain at most two novel alleles per V gene. This guards against over-splitting caused by low-frequency artefacts, sequencing noise, or residual SHM that can mimic additional variants. we prioritize candidates by those with the higher `novel_imgt_count`.

`novel_imgt_count`: The number of times the sequence novel_imgt is found in the input data. Considers the subsequence of novel_imgt in the pos_range.

*Repeated-read / recurrent SNP-pattern filter*
We remove support from sequences that show repeated instances of the same novel-SNP motif inside a single read. These patterns are a common source of false positives from sequecing errors. 

```{r clean-up}

if (!inherits(novel, "try-error") && nrow(novel) > 0) {
  novel <- tigger::selectNovel(novel) %>%
    dplyr::distinct(novel_imgt, .keep_all = TRUE) %>%
    dplyr::filter(!is.na(novel_imgt), nt_substitutions != "") %>%
    dplyr::mutate(gene = alakazam::getGene(germline_call, strip_d = FALSE)) %>%
    dplyr::group_by(gene) %>%
    dplyr::top_n(n = 2, wt = novel_imgt_count)

  snp_xxxx <- check_repeated_snp_patterns(novel)
  novel <- novel[!snp_xxxx, ]
  novel <- novel[!duplicated(novel[["polymorphism_call"]]), ]
}

```

# Generate novel reference database

We append the high-confidence novel alleles to the existing germline reference set, producing an updated “novel” reference database. Each novel allele retains its reference name plus a suffix identifying its novel variant

```{r build-reference}

if (!inherits(novel, "try-error") && nrow(novel) > 0) {
  # Build novel reference database
  novel_germline <- setNames(gsub("-", ".", novel[["novel_imgt"]], fixed = TRUE), novel[["polymorphism_call"]])
  germline_db <- c(germline_db, novel_germline)
}

```

# Reassign sequences to novel alleles (optional)

If enabled, sequences are re-aligned to the updated germline database to include the novel alleles in the repertoire annotations.

For re-alignment the `reassignAlleles` function from TIgGER is used. The function utilizes hamming distance on the gapped sequence (i.e `sequence_alignment`), in order to save time, initial gene assignments are preserved and the allele calls are chosen from among those provided in `germline_db`, based on a simple alignment to the sample sequence.

```{r reassign, eval=params$reassign}
if (params$reassign && exists("germline_db") && !all(is.na(germline_db))) {
  # Read rearrangement data with all columns
  db <- read_rearrangement(params$input)

  # Reassign sequences to novel alleles. Only do it if novel alleles were found
  if (nrow(novel) > 0) {
    # TODO: check to add the option of path to aligner, and keep_gene
    # Using default params for now
    db <- reassignAlleles(db,
                          genotype_db = germline_db,
                          v_call = params$v_call,
                          seq = params$seq)
  }
}
```

# Output files

```{r save-results, results='asis'}
# Save novel report
if (!inherits(novel, "try-error") && nrow(novel) > 0) {
  write.table(novel, file = file.path(params$outdir, report_output_fn),
              row.names = FALSE, quote = FALSE, sep = "\t")
  cat("- **Novel alleles report**: [", report_output_fn, "](", report_output_fn, ") - Detailed results table\n")
}

# Save novel reference database
if (!all(is.na(germline_db)) && exists("germline_db")) {
  tigger::writeFasta(germline_db,
                     file = file.path(params$outdir, ref_output_fn))
  cat("- **Reference database**: [", ref_output_fn, "](", ref_output_fn, ") - FASTA file with novel alleles added (if detected)\n")
}

# Save reassigned sequences (if requested)
if (params$reassign && exists("db")) {
  write_rearrangement(db,
                      file = file.path(params$outdir, output_fn))
  cat("- **Reassigned sequences**: [", output_fn, "](", output_fn, ") - Sequences reassigned to novel alleles\n")
}
```


```{r log-end}
cat("\nEND> InferNovelAlleles", file = log_fn, append = TRUE)
if (params$reassign && exists("db")) {
  cat(paste0("\nOUTPUT> ", output_fn), file = log_fn, append = TRUE)
}
cat(paste0("\nOUTPUT> ", basename(ref_output_fn)), file = log_fn, append = TRUE)
if (exists("novel") && !inherits(novel, "try-error") && nrow(novel) > 0) {
  cat(paste0("\nOUTPUT> ", report_output_fn), file = log_fn, append = TRUE)
  cat(paste0("\nNOVEL-ALLELES-DETECTED> ", nrow(novel)), file = log_fn, append = TRUE)
}
```

```{r, child=c('versions.Rmd')}
```