--- 
title: "Immcantation - enchantR"
subtitle: "Clonal Assignment"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr clonal assignment"
output: enchantr::immcantation
params:
   input: 
      label: "`input`: Path to repertoires files"
      input: file
      value: "https://yale.box.com/shared/static/dajikxtvbc3c5yw1o2v69fjtx5bh58r2.tsv"
   imgt_db:
      label: "`imgt_db`: Path to IMGT reference  database."
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   species:
      label: "`species`"
      input: text
      value: "human"        
   force:
      label: "`force`: if clone_id already exists, overwrite (force=TRUE) or skip clone assignment and render the report (force=FALSE)."
      input: logical        
      value: !r FALSE       
   cloneby:
      label: "`cloneby`"
      input: text
      value: "subject_id"    
   singlecell:
      label: "`singlecell`"
      input: text
      value: NULL          
   threshold:
      label: "`threshold`"
      input: numeric
      value: 1
   min_n:
      label: "`min_n` minimum number of observations to sample to run alakazam::estimateAbundance."
      input: numeric
      value: 30
   outputby:
      label: "`outputby` name of the column in `input` that contains sample identifiers that will be used to split the output db." 
      input: text
      value: "sample_id"
   model:
      label: "`model`"
      input: text
      value: "hierarchical" 
   method:
       label: "`method`"
       input: text
       value: "nt"
   linkage:
       label: "`linkage`"
       input: text
       value: "single"
   nboot:
       label: "`nboot`: number of bootstrap realizations to generate in `estimateAbundance`"
       input: numeric
       value: 200               
   outname:
      label: "`outname`"
      input: text
      value: "clonal-assignment"
   nproc:
      label: "`nproc`"
      input: numeric
      value: 1      
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: '`outdir`: Output directory'
      input: text
      value: !r file.path(getwd(),'enchantr')
   date: 
      label: '`date`: Run date'
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo: 
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options: 
  chunk_output_type: console
---

<div class="introduction_paragraph">
Report of clonal assignment, clone size distribution, clonal abundance, diversity and mutation frequency created by Immcantation package <a href="https://enchantr.readthedocs.io/en/stable/" target="_blank">  enchantR</a>. Immcantation package <a href="https://scoper.readthedocs.io/en/stable/" target="_blank">SCOPer</a> is used to assign clones. Package <a href="https://alakazam.readthedocs.io/en/stable/" target="_blank">Alakazam</a> is used for clonal abundance and diversity analysis. Package <a href="https://shazam.readthedocs.io/en/stable/" target="_blank">SHazaM</a> is used for analysis of somatic hypermutation (SHM). 
</div>

```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4, fig.path = "figures/",
                      echo=params$echo,cache=params$cache,
                      warning=FALSE, message=FALSE,
                      eval.after="fig.cap",
                      out_dir=params$outdir,
                      eval.opts = c('eval', 'echo', 'fig.height', 'fig.width'))

# Libraries
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("alakazam"))
suppressPackageStartupMessages(library("shazam"))
suppressPackageStartupMessages(library("scoper"))
suppressPackageStartupMessages(library("dowser"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("ComplexHeatmap"))
suppressPackageStartupMessages(library("plotly"))

if (!dir.exists(params[['outdir']])) {
    dir.create(params[['outdir']], recursive=T)
}

file.copy(params$logo,
          file.path(params$outdir,"assets", "logo.png"),
          recursive=T, overwrite=T)

singlecell <- params$singlecell
```


# Input parameters

```{r input-parameters, results='asis'}
printParams(params)
```
```{r echo=FALSE, results='asis'}
print_table_caption(tag="input", text="Input parameters.")
save(params, file=file.path(params$outdir,"params.RData"))
```

# Read repertoires
<div class="introduction_paragraph">
Tables summarizing number of sequences in the provided repertoires according to different characteristics.
</div>

```{r}
# Read repertoire
db <- readInput(params[['input']], col_select = NULL)

if (params$species =="auto") {
    if ("species" %in% colnames(db)) {
        species <- unique(db[['species']])
        if (length(species)>1) {
            stop("Multiple species detected. Only one allowed.")
        }
    } else  {
        stop("Can't detect species. The column `species` does not exist in `db`.")
    }
} else {
    species <- params$species
}

# Check for single cell label
if (!is.null(singlecell)) {
    if (singlecell %in% colnames(db)) {
         db[[ singlecell ]] <- as.logical(db[[ singlecell ]] )   
    } else {
        stop("`",singlecell, "` is not a valid field in `db`.")
    }
} else {
   singlecell <- "single_cell"
   db[[singlecell]] <- F
   if ("cell_id" %in% colnames(db)) {
       message("Setting `singlecell` using `cell_id`.") 
       db[[singlecell]][!db[['cell_id']] %in% c(NA, '')] <- T
   }

}

if (!"locus" %in% colnames(db)) {
    db[['locus']] <- getLocus(db[['v_call']])
}
heavy_chains <- isHeavyChain(db[['locus']])

if ("clone_id" %in% colnames(db)) {
    if (params$force) {
        # Reset if force
        warning("Overwritting clone_id.")
        db$clone_id <- NULL
    }   
    # Reset always if clone_id exists
    if ("clone_size_count" %in% colnames(db)) {    
        warning("Overwritting clone_size_count.")    
        db$clone_size_count <- NULL
    }
    if ("clone_size_freq" %in% colnames(db)) {    
        warning("Overwritting clone_size_freq.")    
        db$clone_size_freq <- NULL
    }
}

subject_levels <- str_sort(unique(db$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(db$sample_id),  numeric = TRUE)
db <- db %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id) 
```

```{r}
input_size <- nrow(db)
input_sizes <- db %>%
    group_by(!!!rlang::syms(unique(c("input_file")))) %>%
    summarize(input_size=n())

# Track input size of the expected output groups, for the
# end of report summary
input_sizes_byoutput <- db %>%
    group_by(!!!rlang::syms(unique(params$outputby))) %>%
    mutate(
        num_input_files = length(unique(input_file)),
        input_files = paste(unique(input_file), collapse=",")
    ) %>%
    ungroup() %>%
    group_by(!!!rlang::syms(unique(c(params$outputby, "input_file", "num_input_files", "input_files")))) %>%
    summarize(
        input_size = n()
    )

```


Number of sequences loaded: `r input_size`. Number of heavy chain sequences loaded: `r sum(heavy_chains, na.rm=T)`.

```{r inputsamplesumary, results='asis', eval="tissue" %in% colnames(db)}
input_samples_summary <- db %>%
   group_by(sample_id, subject_id, tissue) %>%
   summarize(size = n()) 

subject_levels <- str_sort(unique(input_samples_summary$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(input_samples_summary$sample_id),  numeric = TRUE)
input_samples_summary <- input_samples_summary %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id,tissue) %>%
  mutate(
    subject_id = as.character(subject_id),
    sample_id  = as.character(sample_id)
  )

eetable(input_samples_summary)$table
```
```{r echo=FALSE, results='asis', eval="tissue" %in% colnames(db)}
print_table_caption(tag="input_samples_summary", text="Input samples summary.")
```


## Sequences per locus

```{r inputlocussumary, results='asis'}
input_locus_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "locus")))) %>%
   summarize(n=n(), .groups="drop") %>%
   pivot_wider(names_from=locus, values_from=n) %>%
   rowwise() %>%
   mutate(Total = sum(!!!rlang::syms(unique(db[['locus']]))))

total <- data.frame(list("sample_id"="Total", 
                    t(input_locus_summary %>%
   select(!!!rlang::syms(c(unique(db[['locus']]), "Total"))) %>%
   colSums(na.rm = T))))

input_locus_summary <- bind_rows(input_locus_summary, total)

subject_levels <- str_sort(unique(input_locus_summary$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(input_locus_summary$sample_id),  numeric = TRUE)
input_locus_summary <- input_locus_summary %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id) %>%
  mutate(
    subject_id = as.character(subject_id),
    sample_id  = as.character(sample_id)
  )

tab_caption <- paste0("Input data. Number of sequences in each ", 
                   paste("sample", params$cloneby, sep=", "),
                   " and locus."
      )
eetable(input_locus_summary)$table
```
```{r echo=FALSE,results='asis'}
print_table_caption(tag="input_locus_summary", text=tab_caption)
```


```{r eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db) , results='asis'}
cat("## Sequences per constant region\n")
```

```{r input-cgene-summary, eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db) , results='asis'}

# Create c_gene column only for plotting (will be removed before storing the df again)
db <- db %>% filter(locus %in% c("IGH", "IGK", "IGL")) %>%
    mutate(c_gene = alakazam::getGene(c_call, first=TRUE))

input_cgene_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "c_gene")))) %>%
   summarize(n=n(), .groups="drop") %>%
       pivot_wider(names_from=c_gene, values_from=n)

subject_levels <- str_sort(unique(input_cgene_summary$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(input_cgene_summary$sample_id),  numeric = TRUE)
input_cgene_summary <- input_cgene_summary %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id)

tab_caption <- paste0("Input data. Number of sequences in each ", 
                   paste("sample", params$cloneby, sep=", "),
                   " and c_gene"
      )
eetable(input_cgene_summary)$table
```

```{r echo=FALSE, eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db), results='asis'}
print_table_caption(tag="input_cgene_summary", text=tab_caption)
```

# Clonal assignment

<div class="introduction_paragraph">
Clonal assignment is performed with the function `hierarchicalClones` from package <a href="https://scoper.readthedocs.io/en/stable/" target="_blank">SCOPer</a>. This approach clusters B or T cell receptor sequences based on junction region sequence similarity within partitions that share the same V gene, J gene, and junction length, allowing for ambiguous V or J gene annotations. By default, 'subject_id' is used for grouping samples before assigning clones. 
</div>

```{r, eval="clone_id" %in% colnames(db), results='asis', echo=FALSE}
cat("\n\nThe input data contains clonal assignments, and are being reused.")
```

```{r, child=c('run_scoper.Rmd'), eval=!"clone_id" %in% colnames(db)}
```

# Clone size distribution

<div class="introduction_paragraph">
The clone sizes table is available <a href=tables/clone_sizes_table.tsv target="_blank">here</a>. In real datasets, most clones contain only one unique sequence as they represent naive cells that have not undergone clonal expansion (also termed singletons). Straight sequence count as a measure of the size of the clones is not the best measure for clonal size comparison between samples due to possible disproportionate sampling. The clonal abundance and diversity computations (see sections below) address this issue by bootstrapping (i.e. re-sampling multiple times) the sample sequences to the same number of sequences. 
</div>

<div class="introduction_paragraph">
Description of terms:

* `clone_size_count`: Clone size as sequence counts. In a sample (`sample_id`), the number of heavy chain sequences with the same `clone_id`.

* `clone_size_freq`: Clone size as percent of the repertoire. `clone_size_count` divided by the number of heavy chain sequences in the sample (`sample_id`).
</div>


## Number of clones 
<div class="introduction_paragraph">
Summary of the number of clones based on heavy chain sequences, and clone size per sample. Singletons (clones with just one sequence) are included in the table. 
</div>

```{r clone_sizes_table}
# Add clone_size 
clone_sizes <- countClones(
            db %>% filter(isHeavyChain(locus)), # Keep heavy chains only
            groups=unique(c("sample_id", params$cloneby)))

subject_levels <- str_sort(unique(clone_sizes$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(clone_sizes$sample_id),  numeric = TRUE)
clone_sizes <- clone_sizes %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id)

```


```{r eval=any(isHeavyChain(db$locus))}
tmp <- eetable(clone_sizes, 
               caption=NULL, 
               outdir=params$outdir, file="clone_sizes_table")

db <- db %>%
   left_join(clone_sizes) %>%
   rename(
      clone_size_count = seq_count,
      clone_size_freq = seq_freq
   ) %>% 
   mutate_at(vars(starts_with("clone_size")), round,2)
```

```{r, results='asis', eval=any(isHeavyChain(db$locus))}
num_clones_table <- db %>%
   filter(isHeavyChain(locus)) %>% # Keep heavy chains only
   group_by(sample_id) %>%
   mutate(sequences=n()) %>%
   group_by(sample_id) %>%
   mutate(
      number_of_clones=length(unique(clone_id)),
   ) %>%
   group_by(!!!rlang::syms(unique(c("sample_id","sequences", params$cloneby, "number_of_clones")))) %>%
   summarize_at(vars(starts_with("clone_size")), list("min"=min, "median"=median, "max"=max)) %>%
   mutate_at(vars(starts_with("clone_size")), round, 2)

subject_levels <- str_sort(unique(num_clones_table$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(num_clones_table$sample_id),  numeric = TRUE)
num_clones_table <- num_clones_table %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id) %>%
  mutate(
    subject_id = as.character(subject_id),
    sample_id  = as.character(sample_id)
  )

tab_caption <- "Summary of the number of clones, and clone size, per sample. Includes singletons (clone_size == 1)."
tab <- eetable(num_clones_table, caption=tab_caption, outdir=params$outdir, file="num_clones_table")
tab$table
```
```{r echo=FALSE, results='asis', eval=any(isHeavyChain(db$locus))}
print_table_caption(tag="num_clones_table", text=tab$caption)
```

## Clone size distribution 

<div class="introduction_paragraph">
Plot of clone size distribution. Clone size is measured as number of heavy chain sequences belonging to the same clone. In most cases, the majority of clones are singletons (clone size = 1). To improve readability for non-singleton clones, we adopted two approaches and provided two corresponding plots. In the first plot, the y-axis (clone counts) was transformed to a logarithmic scale (base 10). In the second plot, the y-axis represents the cumulative fraction of clone counts, also known in statistics as the cumulative distribution function (CDF).
</div>

```{r clone-size-plot, eval=max(clone_sizes[['seq_count']])>1, fig.width=8, fig.height=min(50,max(4,0.7*length(unique(db[['sample_id']])))), fig.cap=caption}
caption <- "Clone size distribution."

clone_sizes_summary <- clone_sizes %>%
  count(sample_id, seq_count, name = "count")

clone_size_plot<-ggplot(clone_sizes_summary, aes(x=seq_count, y=count,
                                         color=sample_id, 
                                         fill=sample_id, 
                                         text = paste("Clone size: ", seq_count, 
                                                                 "<br>Clone count: ", count, 
                                                                 "<br>sample_id: ", sample_id))) +
                geom_bar(stat = "identity") + 
                scale_y_log10()+
                theme_enchantr() +
                facet_wrap(~sample_id, scales = "free_y", ncol=3) +
                xlab("Clone size (Sequences per clone)") +
                ylab("Clone counts") + 
  theme(text = element_text(family = "Helvetica"))

clone_size_plot <- eeplot(clone_size_plot, 
                  outdir=params$outdir, 
                  file=knitr::opts_current$get('clone-size-plot'),
                  caption=caption)
                
ggplotly(clone_size_plot + theme(panel.spacing=unit(2, 'lines'), legend.position="right"),tooltip = "text")  %>%
  layout(font = list(family = "Helvetica"))
caption <- clone_size_plot$enchantr$html_caption
```


```{r clone-size-cdf-plot, eval=max(clone_sizes[['seq_count']])>1, fig.width=8, fig.height=min(50,max(4,0.7*length(unique(db[['sample_id']])))), fig.cap=caption}
caption <- "CDF plot of clone size distribution."

# Step : Calculate count and CDF
clone_sizes_summary <- clone_sizes %>%
  count(sample_id, seq_count, name = "count") %>%
  group_by(sample_id) %>%
  arrange(seq_count) %>%
  mutate(
    cum_count = cumsum(count),
    total = sum(count),
    cdf = cum_count / total,
    text = paste("Clone size:", seq_count,
                 "<br>Clone count:", count,
                 "<br>sample_id:", sample_id,
                 "<br>CDF:", signif(cdf, 3))
  )

# Step 2: CDF plot with line and points
clone_size_cdf_plot <- ggplot(clone_sizes_summary, aes(x = seq_count, y = cdf,
                                                       color = sample_id,
                                                       group = sample_id,
                                                       text = text)) +
  geom_line() +
  geom_point(size = 1, alpha = 0.9)+
  scale_y_log10() +
  scale_x_log10() +  # Optional: log-transform x-axis for clarity
  theme_enchantr() +
  facet_wrap(~sample_id, scales = "free_y", ncol = 3) +
  xlab("Clone size (Sequences per clone)") +
  ylab("Cumulative Fraction") + 
  theme(text = element_text(family = "Helvetica"))

clone_size_cdf_plot <- eeplot(clone_size_cdf_plot, 
                  outdir=params$outdir, 
                  file=knitr::opts_current$get('clone-size-cdf-plot'),
                  caption=caption)

# Step 3: Convert to interactive plot
ggplotly(clone_size_cdf_plot + 
           theme(panel.spacing = unit(2, 'lines'), legend.position = "right"),
         tooltip = "text") %>%
  layout(font = list(family = "Helvetica"))


caption <- clone_size_cdf_plot$enchantr$html_caption
```

```{r, eval=max(clone_sizes[['seq_count']])<=1, results='asis'}
cat("Only singletons detected: there aren't clones of size>1.\n\n")
```

# Clonal abundance {#clonal_abundance}

<div class="introduction_paragraph">
Clonal abundance is the size of each clone as a fraction of the entire repertoire. To correct for the different number of sequences in each of the samples, Function `estimateAbundance` from package <a href="https://alakazam.readthedocs.io/en/stable/" target="_blank">Alakazam</a> estimates the clonal abundance distribution along with confidence intervals on these clone sizes using bootstrapping. `r params$nboot` random bootstrap samples were taken, with size the number of sequences in the sample with less sequences (N). The y-axis shows the clone abundance (i.e., the size as a percent of the repertoire) and the x-axis is a rank of each clone, where the rank is sorted by clone size from larger (rank 1, left) to smaller (right). The shaded areas are confidence intervals.
</div>

```{r}
# set empty default abundance
a <- new("AbundanceCurve")
```

```{r abundancebysample, results='asis', eval=any(isHeavyChain(db$locus))}
# calculate the rank-abundance curve
a <- estimateAbundance(db %>% filter(isHeavyChain(locus)),
                       group = "sample_id", min_n = params$min_n)
if (nrow(a@abundance)==0) {
    cat("\nAll groups failed to pass the threshold min_n=",params$min_n,". Skipping clonal abundance report.\n\n")
}
```

```{r, child=c('abundance.Rmd'), eval=nrow(a@abundance)>0}
```

```{r eval=any(db$locus %in% c("IGH", "IGL", "IGK")), results='asis'}
cat("# Mutation frequency\n")
```

<div class="introduction_paragraph">
Somatic Hypermutation frequency (mu_freq) is determined by comparing a sequence to the germline reference. Function `observedMutations` from package <a href="https://shazam.readthedocs.io/en/stable/" target="_blank">SHazaM</a> is used to calculate the observed number of mutations for each entire sequence. 
</div>


```{r mutation_frequency, eval=any(db$locus %in% c("IGH", "IGL", "IGK")) & !("mu_freq" %in% colnames(db)), results='asis'}
db <- shazam::observedMutations(db,
                                sequenceColumn = "sequence_alignment",
                                germlineColumn = "germline_alignment_d_mask",
                                frequency = T,
                                combine = T,
                                nproc = params$nproc)
```

```{r eval=TRUE, echo=F}
getPrettyWidth <- function(nfacets, ncols, min_w=5, max_w=7) {
    ncols <- min(nfacets, ncols)
    min(max_w, max(ncols*2, min_w))
}

getPrettyHeight <- function(nfacets, ncols, min_h=5, max_h=40) {
    nrows <- ceiling(nfacets/ncols)
    min(max_h, max(nrows*2, min_h))
}
```

```{r eval=any(db$locus %in% c("IGH", "IGL", "IGK")), results='asis'}
cat("Showing mutation frequency per c_gene only if the `c_call` column is present.\n")
cat("The mutation frequency per sequence is stored in the final dataframes in the `mu_freq` column.\n")
```


```{r mufreqSample, eval=any(db$locus %in% c("IGH", "IGL", "IGK")) & "c_call" %in% colnames(db), fig.width=getPrettyWidth(length(unique(db[['sample_id']])), ncols=3), fig.height=getPrettyHeight(length(unique(db[['sample_id']])),ncols=3), fig.cap=caption}
db_cgene <- db %>% filter(!is.na(c_gene), !(is.na(mu_freq)))
db_cgene$mu_freq <- as.numeric(db_cgene$mu_freq)

subject_levels <- str_sort(unique(db_cgene$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(db_cgene$sample_id),  numeric = TRUE)
db_cgene <- db_cgene %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id) 


if (nrow(db_cgene) > 0) {
    mufreqSample <- ggplot(db_cgene, aes(x=c_gene, y=mu_freq, fill=c_gene)) + 
                    geom_boxplot() +
                    facet_wrap(~ subject_id + sample_id, ncol=3, scales = "free") +
                    theme_enchantr() +
                    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    mufreqSample <- eeplot(mufreqSample, 
                        outdir=params$outdir, 
                        file='mutation_frequency_sample',
                        caption=paste0("Mutation frequency per C gene per `sample_id`."),
                        mufreqSample)
    print(mufreqSample +
        theme(legend.position = "none"))
    caption <- mufreqSample$enchantr$html_caption  
}

```

```{r mufreqSampleLocus, eval=any(db$locus %in% c("IGL", "IGK")), fig.width=getPrettyWidth(length(unique(db[['sample_id']])), ncols=3), fig.height=getPrettyHeight(length(unique(db[['sample_id']])),ncols=3), fig.cap=caption}

db_locus <- db %>% filter(!is.na(locus), !is.na(mu_freq))
db_locus$mu_freq <- as.numeric(db_locus$mu_freq)

subject_levels <- str_sort(unique(db_locus$subject_id), numeric = TRUE)
sample_levels  <- str_sort(unique(db_locus$sample_id),  numeric = TRUE)
db_locus <- db_locus %>%
  mutate(
    subject_id = factor(subject_id, levels = subject_levels),
    sample_id  = factor(sample_id,  levels = sample_levels)
  ) %>%
  arrange(subject_id, sample_id) 

if (nrow(db_locus) > 0) {
    mufreqSampleLocus <- ggplot(db_locus, aes(x=locus, y=mu_freq, fill=locus)) + 
                    geom_boxplot() +
                    facet_wrap(~ subject_id + sample_id, ncol=3, scales = "free") +
                    theme_enchantr() +
                    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    mufreqSampleLocus <- eeplot(mufreqSampleLocus, 
                        outdir=params$outdir, 
                        file='mutation_frequency_sample_locus',
                        caption=paste0("Mutation frequency per locus per `sample_id`."),
                        mufreqSampleLocus)
    print(mufreqSampleLocus +
        theme(legend.position = "none"))
    caption <- mufreqSampleLocus$enchantr$html_caption 
}

```

# Final repertoires and tables

Summary tables:

```{r, results='asis', echo=FALSE}
cat("- [Summary of number of clones](tables/num_clones_table.tsv)\n", sep="")
cat("- [Clone sizes table](tables/clone_sizes_table.tsv)\n", sep="")
```
```{r, results='asis', echo=FALSE, eval=nrow(a@abundance)>0}
cat("- [Clonal abundance](tables/clonal_abundance.tsv)\n", sep="")
cat("- [Clonal diversity](tables/clonal_diversity.tsv)\n", sep="")
```

Find your final repertoires here:

```{r fn, results='asis', echo=F}
# pass <- db[['collapse_pass']]

if (!is.null(params$outname)) {
    output_fn <- paste0(params$outname,"_clone-pass.tsv")
} else {
    output_fn <- sub(".tsv$", "_clone-pass.tsv", basename(params$input))
}

if (!is.null(params$log)) {
    log_fn <- file.path(params$outdir,paste0(params$log,".txt"))
} else {
    log_fn <- file.path(params$outdir,sub("_clone-pass.tsv$", "_command_log.txt", basename(output_fn)))
}
```

```{r dbpass, results='asis', echo=FALSE}

# Remove unnecessary columns
db <- db %>% select(-clone_size_count, -clone_size_freq, -any_of(c("c_gene")))

# collpase in case there are multiple rows
# due to multiple input files
input_sizes_byoutput <- input_sizes_byoutput %>%
    select(-input_files) %>%
    ungroup() %>%
    group_by(!!!rlang::syms(params$outputby)) %>%
    summarise(across(everything(), function(x) {paste(x, collapse = ",")} ))
    
output_groups <- db %>%
    ungroup() %>%
    select(!!!rlang::syms(c(params$outputby, "input_file"))) %>%
    group_by(!!!rlang::syms(params$outputby)) %>%
    summarize(output_size = n()) %>%
    right_join(input_sizes_byoutput) %>%
    select(input_size, output_size, everything()) 
    
   
output_files <- c()
output_sizes <- c()
input_sizes <- c()
dir.create(file.path(params$outdir,"repertoires")) 
for (i in 1:nrow(output_groups))  {
    group_db <- left_join( output_groups[i,params$outputby,drop=F],
                           db,
                           by=c(params$outputby))
    
    group_db_label <- gsub("^id_","",makeLabel(group_db, fields= params$outputby))
    group_fn <- file.path(params$outdir,"repertoires",paste0(group_db_label,"_",output_fn))
    output_files <- c(output_files, group_fn)
    output_sizes <- c(output_sizes, nrow(group_db))
    i_input_size <- sum(as.numeric(strsplit(output_groups[["input_size"]][i],",")[[1]]))
    input_sizes <- c(input_sizes, i_input_size)
    write_rearrangement(group_db %>% select(-input_file), file=group_fn)
    cat("- [",group_db_label,"](",file.path("repertoires",basename(group_fn)),")\n", sep="")
}
```

```{r log, echo=FALSE}
cat("START> ClonePass", file=log_fn, append=F)
for (i in 1:nrow(output_groups)) {
    cat(paste0("\nFILE",i,"> ",basename(output_groups[['input_file']][i])), file=log_fn, append=T)
    cat(paste0("\nOUTPUT",i,"> ",basename(output_files[i])), file=log_fn, append=T)
    cat(paste0("\nPASS",i,"> ",output_sizes[i]), file=log_fn, append=T)
    cat(paste0("\nFAIL",i,"> ",input_sizes[i]-output_sizes[i]), file=log_fn, append=T)    
}
```

```{r, child=c('versions.Rmd')}
```