Clonal assignment performed with `scoper::hierarchicalClones`, version `r v <- packageVersion("scoper"); v` within `r paste(params$cloneby, collapse=", ")`.

To know more details about the method, visit the documentation website [https://scoper.readthedocs.io/en/](https://scoper.readthedocs.io/en/`r v`/topics/hierarchicalClones/)

```{r, warning=TRUE}
if (sum(heavy_chains)>0) {
    if (params$model == "hierarchical") {
        if (all(db[[ singlecell ]] == T )) {
            cell_id <- 'cell_id'
        } else {
            db_light <- db %>% filter(!isHeavyChain(locus))
            cell_id <- NULL
            if (all(c(T,F) %in% db[[ singlecell ]])) {
                warning("Mix of single and bulk data. Setting cell_id=`NULL`.")
            }
        }
        db <- hierarchicalClones(db, 
                                 params$threshold, 
                                 method=params$method,
                                 linkage=params$linkage, 
                                 normalize="len",
                                 junction="junction", 
                                 v_call="v_call", j_call="j_call", 
                                 clone="clone_id", 
                                 fields=params$cloneby,
                                 cell_id=cell_id, 
                                 locus="locus", 
                                 only_heavy=TRUE, 
                                 split_light=TRUE,
                                 first=FALSE, 
                                 cdr3=FALSE, mod3=FALSE, 
                                 max_n=0, nproc=params$nproc,
                                 verbose=FALSE, log=NULL,
                                 summarize_clones=FALSE) 
    } else {
        stop("Unsuported model requested. Supported models: hierarchical")
    }
} else {
    warning("No heavy chain sequences found.")
    db$clone_id <- NA
}
```

`r nrow(db)` sequences passed the clonal assignment step 
and `r input_size - nrow(db)` were removed. `r sum(is.na(db[['clone_id']]))` sequences
have `clone_id==NA`.

## Create germlines

```{r}
dowser_v <- packageVersion("dowser")
```


Identification of the V(D)J germline sequences from which each of the observed 
sequences is derived is performed with `dowser::createGermlines`, version `r dowser_v`.
These reconstructed germlines will be used in downstream analysis to infer somatic 
mutations and reconstruct lineages. 

`dowser::createGermlines` takes the alignment information 
in the rearrangement file as well as the reference database used by the 
alignment software and generates a germline sequence for each individual observed sequence.
Because clonal relations have already been inferred, the function assigns the 
same germline to all sequences belonging to the same clone. 

Two types of germlines
are created, `germline_alignment` and `germline_alignment_d_mask`. The last one
has the D region masked, meaning that all nucleotides in the N/P and D-segments
are replaced with N's. This is often done because the germline base calls from 
this region are unreliable for B cell receptor alignments.

Documentation for `dowser::createGermlines` is available here: [https://dowser.readthedocs.io/en/latest/topics/createGermlines](https://dowser.readthedocs.io/en/latest/topics/createGermlines/).

```{r}
pre_germ_size <- nrow(db)
db[['tmp_nrow']] <- 1:nrow(db)

# download and unzip if needed
imgt_db <- prepareIMGT(params$imgt_db)

references <- dowser::readIMGT(file.path(imgt_db,
                                species, "vdj"),
                               quiet=TRUE)

# tmp fix (add unique sequence id ) for
# Error in dowser::createGermlines(db_sp, references, locus = "locus",  :
#   Sequence IDs are not unique!
dup_ids <- db %>%
    ungroup() %>%
    group_by(!!!rlang::syms(c(params$cloneby))) %>%
    mutate(dup=duplicated(sequence_id)) %>%
    pull(dup) %>% any()

if (dup_ids) {
    db <- db %>%
        mutate(sequence_id = paste(sample_id, sequence_id, sep = '_'))
}

is_na_clone <- is.na(db[["clone_id"]])
not_na_clone_db <- data.frame()
na_clone_db <- data.frame()
if (any(is_na_clone)) {
    na_clone_db <- dowser::createGermlines(
        db[is_na_clone,,drop=FALSE],
        references,
        locus = "locus",
        nproc = params$nproc,
        seq = "sequence_alignment",
        v_call = "v_call",
        d_call = "d_call",
        j_call = "j_call",
        amino_acid = FALSE,
        id = "sequence_id",
        clone = "tmp_nrow",
        v_germ_start = "v_germline_start",
        v_germ_end = "v_germline_end",
        v_germ_length = "v_germline_length",
        d_germ_start = "d_germline_start",
        d_germ_end = "d_germline_end",
        d_germ_length = "d_germline_length",
        j_germ_start = "j_germline_start",
        j_germ_end = "j_germline_end",
        j_germ_length = "j_germline_length",
        np1_length = "np1_length",
        np2_length = "np2_length",
        na.rm = TRUE,
        fields = params$cloneby
    )    
}

if (any(!is_na_clone)) {
    not_na_clone_db <- dowser::createGermlines(
        db[!is_na_clone,,drop=FALSE],
        references,
        locus = "locus",
        nproc = params$nproc,
        seq = "sequence_alignment",
        v_call = "v_call",
        d_call = "d_call",
        j_call = "j_call",
        amino_acid = FALSE,
        id = "sequence_id",
        clone = "clone_id",
        v_germ_start = "v_germline_start",
        v_germ_end = "v_germline_end",
        v_germ_length = "v_germline_length",
        d_germ_start = "d_germline_start",
        d_germ_end = "d_germline_end",
        d_germ_length = "d_germline_length",
        j_germ_start = "j_germline_start",
        j_germ_end = "j_germline_end",
        j_germ_length = "j_germline_length",
        np1_length = "np1_length",
        np2_length = "np2_length",
        na.rm = TRUE,
        fields = params$cloneby
    )
}

db <- bind_rows(not_na_clone_db, na_clone_db) %>%
    arrange(tmp_nrow) %>%
    select(-tmp_nrow)

any_germ_fail <- pre_germ_size > nrow(db)
```

`r nrow(db)` sequences passed the germline reconstruction step
and `r pre_germ_size - nrow(db)` failed.

```{r eval=(any_germ_fail & !is.null(cell_id)) }
# Remove cells that after createGermlines
# have only light chains
db <- findLightOnlyCells(db, 
                         sample_id = "sample_id", cell_id = cell_id,
                         locus = "locus", fields = NULL)
num_light_only <- sum(db[["light_only_cell"]], na.rm = T)
if (num_light_only > 0) {
    # Remove the cells
    db <- db %>% dplyr::filter(!light_only_cell)
    cat(paste0("Removed ",num_light_only," sequences in light chain only cells."))
}
db[["light_only_cell"]] <- NULL
```

```{r eval=any(db[[ singlecell ]] == F), results='asis'}
# Add again light chains before observed mutations when mixture of sc + bulk or only bulk
if (nrow(db_light) > 0){
    db <- bind_rows(db, db_light)
}

```