
```{r abundanceAnnotation}
# annotate
a@abundance <- a@abundance %>%
   left_join( db %>% 
                 select(any_of(c(unique(c("sample_id", 
                                                "subject_id",
                                                params$cloneby)))
                                       )
                  ) %>%
                 distinct(),
              
              )
p <- plotAbundanceCurve(a, annotate="depth", silent = T)
```

```{r clonalAbundanceTable, results='asis'}
tab <- eetable(a@abundance, file = "clonal_abundance", 
               outdir=params$outdir, show_max=10,
               caption="Example 10 lines of the clonal abundance file.")
tab$table %>%
        DT::formatRound(columns=c("p","p_sd","lower","upper"),digits=3)
```
```{r echo=FALSE, results='asis'}
print_table_caption("clonal_abundance", tab$caption)
```

## Abundance plot by sample

```{r eval=TRUE, echo=F}
getPrettyWidth <- function(nfacets, ncols, min_w=5, max_w=7) {
    ncols <- min(nfacets, ncols)
    min(max_w, max(ncols*2, min_w))
}

getPrettyHeight <- function(nfacets, ncols, min_h=5, max_h=20) {
    nrows <- ceiling(nfacets/ncols)
    min(max_h, max(nrows*2, min_h))
}
```


```{r abundanceSample, eval=TRUE, fig.width=getPrettyWidth(length(unique(db[['sample_id']])), ncols=3), fig.height=getPrettyHeight(length(unique(db[['sample_id']])),ncols=3), fig.cap=caption}
abundanceSample <- p + facet_wrap(~ subject_id + sample_id, ncol=3)
abundanceSample <- eeplot(abundanceSample, 
                     outdir=params$outdir, 
                     file=knitr::opts_current$get('abundanceSample'),
                     caption=paste0("Clonal abundance plot per `sample_id`. Clonal abundance was calculated with a sample of N=",
                      a@n,
                      " sequences with ",
                      a@nboot,
                      " boostrapping repetitions."),
                     a)
abundanceSample +
    theme(legend.position = "none")
caption <- abundanceSample$enchantr$html_caption  
```

# Diversity

The clonal abundance distribution can be characterized using diversity statistics. Diversity scores (D) are calculated using the generalized diversity index (Hill numbers), which covers many different measures of diversity in a single function with a single varying parameter, the diversity order q.

The function alphaDiversity resamples the sequences (`r params$nboot` random bootstrapping events, with the number of sequences in the sample with less sequences (N)) and calculates diversity scores (D) over a interval of diversity orders (q). The diversity (D) is shown on the y-axis and the x-axis is the parameter q. - q = 0 corresponds to Species Richness - q = 1 corresponds to Shannon Entropy - q = 2 corresponds to Simpson Index

Inspection of this figure is useful to determine whether any difference in diversity between two repertoires depends on the statistic used or if it is a universal property. 


The clonal diversity $D$ of the repertoire was calculated according to the general formula of Hill Diversity
numbers:

$$
\begin{aligned}
    ^{q}D = \left( \sum_{i=1}^Rp_i^q \right)^{1/(1-q)}
\end{aligned}
$$

where:

* $p_i$ is the proportion of unique sequences belonging to clone $i$.
* $q$ are the values of the different diversity numbers.
* $R$ is the Richness, the number of different clones in the sample.

At $q=1$ the function is undefined and the limit to zero equals the exponential of the Shannon Entropy:

$$
\begin{aligned}
    ^{1}D = exp \left(  \sum_{i=1}^Rp_i ln(p_i)  \right)
\end{aligned}
$$

The intuition about the different Hill Diversity values is the following:

* At $q=0$ the diversity index equals the number of clones in the sample.
* At $q=1$ the diversity index is the geometric mean of the clones in the sample,
weighted by their proportion in the sample.
* At $q>1$ more weight is given to the clones with higher proportions in the sample.


## Diversity curves

The following table shows the summary of the diversity calculations per sample.

```{r diversity, results='asis'}
# generate the Hill diversity curve
d <- alphaDiversity(db %>% filter(isHeavyChain(locus)), 
                    group = "sample_id", 
                    min_n=params$min_n)

d@diversity <- d@diversity %>%
   left_join( db %>% 
                 select(any_of(c(unique(c("sample_id", 
                                                "subject_id", 
                                                params$cloneby)))
                                       )
                  ) %>%
                 distinct(),
              
              )

diversitySample <- plotDiversityCurve(d, silent = T, annotate="depth")

tab <- eetable(d@diversity, file = "clonal_diversity", 
               outdir=params$outdir, show_max=10,
               caption="Example 10 lines of the clonal diversity file.")
tab$table %>%
        DT::formatRound(columns=c("q","d","d_sd","d_lower", "d_upper", "e", "e_lower", "e_upper"),digits=3)
```
```{r echo=FALSE,results='asis'}
print_table_caption("clonal_diversity", tab$caption)
```

## Diversity plot by sample

```{r diversitySample, eval=TRUE,  fig.cap=caption, fig.width=getPrettyWidth(length(unique(db[['sample_id']])), ncols=3), fig.height=getPrettyHeight(length(unique(db[['sample_id']])),ncols=3)}
# plot duplicated cells
diversitySample <- diversitySample + 
    geom_vline(xintercept = c(0,1,2), color = "grey50", linetype = "dashed") +
    facet_wrap(~sample_id + subject_id, ncol=3)
diversitySample <- eeplot(diversitySample, 
                     outdir=params$outdir, 
                     file=knitr::opts_current$get('label'),
                     caption=paste0("Clonal diversity per `sample_id`.Clonal diversity was calculated with a sample of N=",
                      d@n,
                      " sequences with ",
                      a@nboot,
                      " boostrapping repetitions."),
                     d)
diversitySample +
    theme(legend.position = "none")
caption <- diversitySample$enchantr$html_caption
```

## Diversity at different `q`

- `q=0` corresponds to Richness (number of different clones in the sample).
- `q=1` corresponds to Shannon diversity.
- `q=2` corresponds to Simpson diversity.

```{r div_at_q, eval=TRUE,  fig.cap=caption, fig.width=5, fig.height=3}
# plot duplicated cells
div_data <- d@diversity %>% dplyr::filter(q==0 | q==1 | q==2)

div_at_q <- ggplot(div_data, aes(x=q, y=d, color=sample_id)) +
    geom_point() + theme_enchantr() +
    scale_x_continuous(breaks= c(0,1,2))
    
div_at_q <- eeplot(div_at_q, 
                     outdir=params$outdir, 
                     file=knitr::opts_current$get('label'),
                     caption="Diversity at different q values.")
div_at_q
caption <- div_at_q$enchantr$html_caption
```
