--- 
title: "Immcantation - enchantR"
subtitle: "Define Clones"
author: ""
date: "Updated: `r date()`"
knit: enchantr::render_book
site: bookdown::bookdown_site
documentclass: book
bibliography: "references.bib"
biblio-style: apalike
link-citations: yes
description: "enchantr define clones"
output: enchantr::immcantation
params:
   input: 
      label: "`input`: Path to repertoires files"
      input: file
      value: "https://yale.box.com/shared/static/dajikxtvbc3c5yw1o2v69fjtx5bh58r2.tsv"
   imgt_db:
      label: "`imgt_db`: Path to IMGT reference  database."
      input: file
      value: "https://raw.githubusercontent.com/nf-core/test-datasets/airrflow/database-cache/imgtdb_base.zip"
   species:
      label: "`species`"
      input: text
      value: "human"        
   force:
      label: "`force`: if clone_id already exists, overwrite (force=TRUE) or skip clone assignment and render the report (force=FALSE)."
      input: logical        
      value: !r FALSE       
   cloneby:
      label: "`cloneby`"
      input: text
      value: "subject_id"    
   singlecell:
      label: "`singlecell`"
      input: text
      value: NULL          
   threshold:
      label: "`threshold`"
      input: numeric
      value: 1
   min_n:
      label: "`min_n` minimum number of observations to sample to run alakazam::estimateAbundance."
      input: numeric
      value: 30
   outputby:
      label: "`outputby` name of the column in `input` that contains sample identifiers that will be used to split the output db." 
      input: text
      value: "sample_id"
   model:
      label: "`model`"
      input: text
      value: "hierarchical" 
   method:
       label: "`method`"
       input: text
       value: "nt"
   linkage:
       label: "`linkage`"
       input: text
       value: "single"
   nboot:
       label: "`nboot`: number of bootstrap realizations to generate in `estimateAbundance`"
       input: numeric
       value: 200               
   outname:
      label: "`outname`"
      input: text
      value: "define-clones"
   nproc:
      label: "`nproc`"
      input: numeric
      value: 1      
   log:
      label: "`log`"
      input: text
      value: "command_log"
   outdir:
      label: '`outdir`: Output directory'
      input: text
      value: !r file.path(getwd(),'enchantr')
   date: 
      label: '`date`: Run date'
      input: date
      value: !r format(Sys.time(), "%Y-%m-%d")
   logo:
      label: "`logo`: Path to report logo"
      input: file
      value: !r file.path("assets", "logo.png")
   logolink:
      label: "`logolink`: URL to be added to the logo"
      input: text
      value: "immcantation.org"
   echo: 
      label: '`echo`: Show code in the report.'
      input: checkbox
      value: TRUE
   cache:
      label: '`cache`: Use cached results'
      input: checkbox
      value: FALSE
editor_options: 
  chunk_output_type: console
---


```{r global-options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4, fig.path = "figures/",
                      echo=params$echo,cache=params$cache,
                      warning=FALSE, message=FALSE,
                      eval.after="fig.cap",
                      out_dir=params$outdir,
                      eval.opts = c('eval', 'echo', 'fig.height', 'fig.width'))

# Libraries
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("airr"))
suppressPackageStartupMessages(library("alakazam"))
suppressPackageStartupMessages(library("shazam"))
suppressPackageStartupMessages(library("scoper"))
suppressPackageStartupMessages(library("dowser"))
suppressPackageStartupMessages(library("enchantr"))
suppressPackageStartupMessages(library("ComplexHeatmap"))
suppressPackageStartupMessages(library("plotly"))

if (!dir.exists(params[['outdir']])) {
    dir.create(params[['outdir']], recursive=T)
}

file.copy(params$logo,
          file.path(params$outdir,"assets", "logo.png"),
          recursive=T, overwrite=T)

singlecell <- params$singlecell
```


# Input parameters

```{r input-parameters, results='asis'}
printParams(params)
```
```{r echo=FALSE, results='asis'}
print_table_caption(tag="input", text="Input parameters.")
save(params, file=file.path(params$outdir,"params.RData"))
```

# Read repertoires

```{r}
# Read repertoire
db <- readInput(params[['input']], col_select = NULL)

if (params$species =="auto") {
    if ("species" %in% colnames(db)) {
        species <- unique(db[['species']])
        if (length(species)>1) {
            stop("Multiple species detected. Only one allowed.")
        }
    } else  {
        stop("Can't detect species. The column `species` does not exist in `db`.")
    }
} else {
    species <- params$species
}

# Check fo single cell label
if (!is.null(singlecell)) {
    if (singlecell %in% colnames(db)) {
         db[[ singlecell ]] <- as.logical(db[[ singlecell ]] )   
    } else {
        stop("`",singlecell, "` is not a valid field in `db`.")
    }
} else {
   singlecell <- "single_cell"
   db[[singlecell]] <- F
   if ("cell_id" %in% colnames(db)) {
       message("Setting `singlecell` using `cell_id`.") 
       db[[singlecell]][!db[['cell_id']] %in% c(NA, '')] <- T
   }

}

if (!"locus" %in% colnames(db)) {
    db[['locus']] <- getLocus(db[['v_call']])
}
heavy_chains <- isHeavyChain(db[['locus']])

if ("clone_id" %in% colnames(db)) {
    if (params$force) {
        # Reset if force
        warning("Overwritting clone_id.")
        db$clone_id <- NULL
    }   
    # Reset always if clone_id exists
    if ("clone_size_count" %in% colnames(db)) {    
        warning("Overwritting clone_size_count.")    
        db$clone_size_count <- NULL
    }
    if ("clone_size_freq" %in% colnames(db)) {    
        warning("Overwritting clone_size_freq.")    
        db$clone_size_freq <- NULL
    }
}
```

```{r}
input_size <- nrow(db)
input_sizes <- db %>%
    group_by(!!!rlang::syms(unique(c("input_file")))) %>%
    summarize(input_size=n())

# Track input size of the expected output groups, for the
# end of report summary
input_sizes_byoutput <- db %>%
    group_by(!!!rlang::syms(unique(params$outputby))) %>%
    mutate(
        num_input_files = length(unique(input_file)),
        input_files = paste(unique(input_file), collapse=",")
    ) %>%
    ungroup() %>%
    group_by(!!!rlang::syms(unique(c(params$outputby, "input_file", "num_input_files", "input_files")))) %>%
    summarize(
        input_size = n()
    )
```


Number of sequences loaded: `r input_size`. Number of heavy chain sequences loaded: `r sum(heavy_chains, na.rm=T)`.

```{r inputsamplesumary, results='asis', eval="tissue" %in% colnames(db)}
input_samples_summary <- db %>%
   group_by(sample_id, subject_id, tissue) %>%
   summarize(size = n()) %>%
   arrange(subject_id)

eetable(input_samples_summary)$table
```
```{r echo=FALSE, results='asis', eval="tissue" %in% colnames(db)}
print_table_caption(tag="input_samples_summary", text="Input samples summary.")
```


## Sequences per locus

```{r inputlocussumary, results='asis'}
input_locus_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "locus")))) %>%
   summarize(n=n(), .groups="drop") %>%
   pivot_wider(names_from=locus, values_from=n) %>%
   rowwise() %>%
   mutate(Total = sum(!!!rlang::syms(unique(db[['locus']]))))

total <- data.frame(list("sample_id"="Total", 
                    t(input_locus_summary %>%
   select(!!!rlang::syms(c(unique(db[['locus']]), "Total"))) %>%
   colSums(na.rm = T))))

input_locus_summary <- bind_rows(input_locus_summary, total)

tab_caption <- paste0("Input data. Number of sequences in each ", 
                   paste("sample", params$cloneby, sep=", "),
                   " and locus."
      )
eetable(input_locus_summary)$table
```
```{r echo=FALSE,results='asis'}
print_table_caption(tag="input_locus_summary", text=tab_caption)
```

```{r eval="c_call" %in% colnames(db), results='asis'}
cat("## Sequences per c_call\n")
```

```{r input-ccall-sumary, eval="c_call" %in% colnames(db), results='asis'}
input_c_call_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "c_call")))) %>%
   summarize(n=n(), .groups="drop") %>%
       pivot_wider(names_from=c_call, values_from=n)

tab_caption <- paste0("Input data. Number of sequences in each ", 
                   paste("sample", params$cloneby, sep=", "),
                   " and c_call"
      )
eetable(input_c_call_summary)$table
```
```{r echo=FALSE, eval="c_call" %in% colnames(db), results='asis'}
print_table_caption(tag="input_c_call_summary", text=tab_caption)
```

```{r eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db) , results='asis'}
cat("## Sequences per isotype\n")
```

```{r input-isotype-sumary, eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db) , results='asis'}
# Create isotype column
db <- db %>%
    mutate(isotype = ifelse(grepl("IGH", locus), alakazam::getGene(c_call, first=TRUE), NA))

input_isotype_summary <- db %>%
   group_by(!!!rlang::syms(unique(c("sample_id",params$cloneby, "isotype")))) %>%
   summarize(n=n(), .groups="drop") %>%
       pivot_wider(names_from=isotype, values_from=n)

tab_caption <- paste0("Input data. Number of sequences in each ", 
                   paste("sample", params$cloneby, sep=", "),
                   " and isotype"
      )
eetable(input_isotype_summary)$table
```

```{r echo=FALSE, eval=any(db$locus %in% c("IGH")) & "c_call" %in% colnames(db), results='asis'}
print_table_caption(tag="input_isotype_summary", text=tab_caption)
```

# Clonal assignment

```{r, eval="clone_id" %in% colnames(db), results='asis', echo=FALSE}
cat("\n\nThe input data contains clonal assingments, and are being reused.")
```

```{r, child=c('run_scoper.Rmd'), eval=!"clone_id" %in% colnames(db)}
```

# Summary of clonal assignment

Description of terms:

* `clone_size_count`: Clone size as sequence counts. In a sample (`sample_id`), the number of heavy chain 
sequences with the same `clone_id`.

* `clone_size_freq`: Clone size as percent of the repertoire. `clone_size_count` divided by the number of heavy chain sequences in the sample (`sample_id`).


## Number of clones (heavy chain, incl. singletons)

```{r clone_sizes_table}
# Add clone_size 
clone_sizes <- countClones(
            db %>% filter(isHeavyChain(locus)), # Keep heavy chains only
            groups=unique(c("sample_id", params$cloneby)))
```


```{r eval=any(isHeavyChain(db$locus))}
tmp <- eetable(clone_sizes, 
               caption=NULL, 
               outdir=params$outdir, file="clone_sizes_table")

db <- db %>%
   left_join(clone_sizes) %>%
   rename(
      clone_size_count = seq_count,
      clone_size_freq = seq_freq
   ) %>% 
   mutate_at(vars(starts_with("clone_size")), round,2)
```

```{r, results='asis', eval=any(isHeavyChain(db$locus))}
num_clones_table <- db %>%
   filter(isHeavyChain(locus)) %>% # Keep heavy chains only
   group_by(sample_id) %>%
   mutate(sequences=n()) %>%
   group_by(sample_id) %>%
   mutate(
      number_of_clones=length(unique(clone_id)),
   ) %>%
   group_by(!!!rlang::syms(unique(c("sample_id","sequences", params$cloneby, "number_of_clones")))) %>%
   summarize_at(vars(starts_with("clone_size")), list("min"=min, "median"=median, "max"=max)) %>%
   mutate_at(vars(starts_with("clone_size")), round, 2)

tab_caption <- "Summary of the number of clones, and clone size, per sample. Includes singletons (clone_size == 1)."
tab <- eetable(num_clones_table, caption=tab_caption, outdir=params$outdir, file="num_clones_table")
tab$table
```
```{r echo=FALSE, results='asis', eval=any(isHeavyChain(db$locus))}
print_table_caption(tag="num_clones_table", text=tab$caption)
```

## Expanded clones within `sample_id` (heavy chain, without singletons)

```{r, results='asis', eval=any(isHeavyChain(db$locus))}
num_clones_table_nosingle <- db %>%
   filter(isHeavyChain(locus)) %>% # Keep heavy chains only
   filter(clone_size_count>1) %>%
   group_by(sample_id) %>%
   mutate(sequences=n()) %>%
   group_by(sample_id) %>%
   mutate(
      number_of_clones=length(unique(clone_id)),
   ) %>%
   group_by(!!!rlang::syms(unique(c("sample_id","sequences", params$cloneby, "number_of_clones")))) %>%
   summarize_at(vars(starts_with("clone_size")), list("min"=min, "median"=median, "max"=max)) %>%
   mutate_at(vars(starts_with("clone_size")), round, 2)

tab_caption <- "Summary of the number of clones of size > 1, and their clone sizes, per sample."
tab <- eetable(num_clones_table_nosingle, caption = tab_caption, outdir=params$outdir, file="num_clones_table_nosingle")
tab$table
```
```{r echo=FALSE,results='asis', eval=any(isHeavyChain(db$locus))}
print_table_caption(tag="num_clones_table_nosingle", text=tab_caption)
```

## Expanded clones within `r paste(params$cloneby,collapse=", ")`

```{r, results='asis', eval=any(isHeavyChain(db$locus)) & "tissue" %in% colnames(db) }
any_tissues <- db %>%
   filter(isHeavyChain(locus)) %>% # Keep heavy chains only
   group_by(!!!rlang::syms(c("clone_id", params$cloneby))) %>%
   summarize(
             clone_size_group_count=n(),
             num_tissues=length(unique(tissue)),
             tissues=paste(sort(unique(tissue)),collapse=" & "), .groups="drop") %>%
   filter(clone_size_group_count>1) %>%
   group_by(!!!rlang::syms(c("tissues", params$cloneby))) %>%
   summarize(num_clones=length(unique(clone_id)))
    
tab_caption <- paste0("Summary of expanded clones in ", paste(params$cloneby, collapse = ", "), " . Subsetting to clones of size > 1 in the cloning group (not the sample).")
eetable(any_tissues, caption=tab_caption)$table
```
```{r echo=FALSE, results='asis', eval=any(isHeavyChain(db$locus))  & "tissue" %in% colnames(db) }
print_table_caption(tag="any_tissues", text=tab_caption)
```

## Clones spanning multiple tissues

```{r, results='asis', eval=any(isHeavyChain(db$locus))  & "tissue" %in% colnames(db) }
multi_tissue <- db %>%
    filter(isHeavyChain(locus)) %>% # Keep heavy chains only
    group_by(!!!rlang::syms(c(params$cloneby, "clone_id"))) %>%
    summarize(num_tissues=length(unique(tissue))) %>%
    filter(num_tissues>1)

multi_tissue <- multi_tissue %>%
   left_join(db %>% filter(isHeavyChain(locus)), 
   by=c("clone_id", params$cloneby)) %>%
   select(!!!rlang::syms(unique(c("clone_id", "tissue", "sample_id", "clone_size_count", "clone_size_freq", params$clone_by, "sequence_id")))) %>%
   group_by(!!!rlang::syms(c("clone_id", "clone_size_count", "clone_size_freq","tissue", "sample_id", params$clone_by))) %>%
   summarize(seq_count=n())

tab_caption <- "Summary of clones spanning more than one tissue."
eetable(multi_tissue, caption=tab_caption)$table
```
```{r echo=FALSE, results='asis',eval=any(isHeavyChain(db$locus))  & "tissue" %in% colnames(db) }
print_table_caption(tag="multi_tissue", text=tab_caption)
```

# Clone size distribution

Find your [clone sizes table here](tables/clone_sizes_table.tsv). Most real datasets, will have most clones of size 1 (one sequence). Straight sequence count as a mesure of the size of the clones is not the best measure to compare clone size between samples. See [Clonal abundance](#clonal_abundance)


```{r clonesize, fig.width=8, fig.height=min(50,max(4,0.7*length(unique(db[['sample_id']])))), fig.cap=caption, eval=any(isHeavyChain(db$locus))}
caption <- "Clone size distribution. Size is measured as number of heavy chain sequences belonging to the same clone."
clone_size_plot <- ggplot(clone_sizes, aes(x=seq_count, color=sample_id, fill=sample_id))+
    geom_bar() + theme_enchantr() +
    facet_wrap(~sample_id, scales = "free_y", ncol=3) +
    xlab("Clone size (Number of sequences per clone)")
clone_size_plot <- eeplot(clone_size_plot, 
       outdir=params$outdir, 
       file=knitr::opts_current$get('clone-size'),
       caption=caption
       )
ggplotly(clone_size_plot + theme(panel.spacing=unit(2, 'lines'), legend.position="right"))
caption <- clone_size_plot$enchantr$html_caption
```

## Subset to clone size > 1

```{r clone-size-atleast2, eval=max(clone_sizes[['seq_count']])>1, fig.width=8, fig.height=min(50,max(4,0.7*length(unique(db[['sample_id']])))), fig.cap=caption}
caption <- "Clone size distribution, excluding singletons (subset to clone size > 1). Size is measured as number of heavy chain sequences belonging to the same clone."
clone_size_atleast2 <- ggplot(clone_sizes %>% filter(seq_count>1), aes(x=seq_count, color=sample_id, fill=sample_id))+
    geom_bar() + theme_enchantr() +
    facet_wrap(~sample_id, scales = "free_y", ncol=3) +
    xlab("Clone size (Sequences per clone)")
clone_size_atleast2 <- eeplot(clone_size_atleast2, 
       outdir=params$outdir, 
       file=knitr::opts_current$get('clone-size_atleast2'),
       caption=caption
       )
ggplotly(clone_size_atleast2 + theme(panel.spacing=unit(2, 'lines'), legend.position="right"))
caption <- clone_size_atleast2$enchantr$html_caption
```

```{r, eval=max(clone_sizes[['seq_count']])<=1, results='asis'}
cat("Only singletons detected: there aren't clones of size>1.\n\n")
```


# Clonal abundance {#clonal_abundance}

```{r}
# set empty default abundance
a <- new("AbundanceCurve")
```

Clonal abundance is the size of each clone (as a fraction of the entire repertoire). To correct for the different number of sequences in each of the samples, `estimateAbundance` estimates the clonal abundance distribution along with confidence intervals on these clone sizes using bootstrapping. `r params$nboot` random bootstrap samples were taken, with size the number of sequences in the sample with less sequences (N). The y-axis shows the clone abundance (i.e., the size as a percent of the repertoire) and the x-axis is a rank of each clone, where the rank is sorted by size from larger (rank 1, left) to smaller (right). The shaded areas are confidence intervals.

```{r abundancebysample, results='asis', eval=any(isHeavyChain(db$locus))}
# calculate the rank-abundance curve
a <- estimateAbundance(db %>% filter(isHeavyChain(locus)),
                       group = "sample_id", min_n = params$min_n)
if (nrow(a@abundance)==0) {
    cat("\nAll groups failed to pass the threshold min_n=",params$min_n,". Skipping clonal abundance report.\n\n")
}
```

```{r, child=c('abundance.Rmd'), eval=nrow(a@abundance)>0}
```


```{r eval=any(db$locus %in% c("IGH", "IGL", "IGK")), results='asis'}
cat("# Mutation frequency\n")
```

```{r mutation_frequency, eval=any(db$locus %in% c("IGH", "IGL", "IGK")), results='asis'}
db <- shazam::observedMutations(db,
                                sequenceColumn = "sequence_alignment",
                                germlineColumn = "germline_alignment_d_mask",
                                frequency = T,
                                combine = T,
                                nproc = params$nproc)
```

```{r eval=TRUE, echo=F}
getPrettyWidth <- function(nfacets, ncols, min_w=5, max_w=7) {
    ncols <- min(nfacets, ncols)
    min(max_w, max(ncols*2, min_w))
}

getPrettyHeight <- function(nfacets, ncols, min_h=5, max_h=12) {
    nrows <- ceiling(nfacets/ncols)
    min(max_h, max(nrows*2, min_h))
}
```

```{r eval=any(db$locus %in% c("IGH", "IGL", "IGK")), results='asis'}
cat("Showing mutation frequency per isotype of heavy chains only if the `c_call` field is present.\n")
cat("The mutation frequency per sequence is stored in the final dataframes in the `mu_freq` column.\n")
```


```{r mufreqSample, eval=any(db$locus %in% c("IGH", "IGL", "IGK")) & "c_call" %in% colnames(db), fig.width=getPrettyWidth(length(unique(db[['sample_id']])), ncols=3), fig.height=getPrettyHeight(length(unique(db[['sample_id']])),ncols=3), fig.cap=caption}

mufreqSample <- ggplot(db %>% filter(!is.na(isotype), locus=="IGH"), aes(x=isotype, y=mu_freq, fill=isotype)) + 
                    geom_boxplot() +
                    facet_wrap(~ subject_id + sample_id, ncol=3) +
                    theme_enchantr() +
                    theme(axis.text.x = element_text(angle = 45, hjust = 1))

mufreqSample <- eeplot(mufreqSample, 
                     outdir=params$outdir, 
                     file='mutation_frequency_sample',
                     caption=paste0("Mutation frequency per isotype per `sample_id`."),
                     mufreqSample)
mufreqSample +
    theme(legend.position = "none")
caption <- mufreqSample$enchantr$html_caption  
```

# Final repertoires and tables

Summary tables:

```{r, results='asis', echo=FALSE}
cat("- [Summary of number of clones](tables/num_clones_table.tsv)\n", sep="")
cat("- [Clone sizes table](tables/clone_sizes_table.tsv)\n", sep="")
```
```{r, results='asis', echo=FALSE, eval=nrow(a@abundance)>0}
cat("- [Clonal abundance](tables/clonal_abundance.tsv)\n", sep="")
cat("- [Clonal diversity](tables/clonal_diversity.tsv)\n", sep="")
```

Find your final repertoires here:

```{r fn, results='asis', echo=F}
# pass <- db[['collapse_pass']]

if (!is.null(params$outname)) {
    output_fn <- paste0(params$outname,"_clone-pass.tsv")
} else {
    output_fn <- sub(".tsv$", "_clone-pass.tsv", basename(params$input))
}

if (!is.null(params$log)) {
    log_fn <- file.path(params$outdir,paste0(params$log,".txt"))
} else {
    log_fn <- file.path(params$outdir,sub("_clone-pass.tsv$", "_command_log.txt", basename(output_fn)))
}
```

```{r dbpass, results='asis', echo=FALSE}

# collpase in case there are multiple rows
# due to multiple input files
input_sizes_byoutput <- input_sizes_byoutput %>%
    select(-input_files) %>%
    ungroup() %>%
    group_by(!!!rlang::syms(params$outputby)) %>%
    summarise(across(everything(), function(x) {paste(x, collapse = ",")} ))
    
output_groups <- db %>%
    ungroup() %>%
    select(!!!rlang::syms(c(params$outputby, "input_file"))) %>%
    group_by(!!!rlang::syms(params$outputby)) %>%
    summarize(output_size = n()) %>%
    right_join(input_sizes_byoutput) %>%
    select(input_size, output_size, everything()) 
    
   
output_files <- c()
output_sizes <- c()
input_sizes <- c()
dir.create(file.path(params$outdir,"repertoires")) 
for (i in 1:nrow(output_groups))  {
    group_db <- left_join( output_groups[i,params$outputby,drop=F],
                           db,
                           by=c(params$outputby))
    
    group_db_label <- gsub("^id_","",makeLabel(group_db, fields= params$outputby))
    group_fn <- file.path(params$outdir,"repertoires",paste0(group_db_label,"_",output_fn))
    output_files <- c(output_files, group_fn)
    output_sizes <- c(output_sizes, nrow(group_db))
    i_input_size <- sum(as.numeric(strsplit(output_groups[["input_size"]][i],",")[[1]]))
    input_sizes <- c(input_sizes, i_input_size)
    write_rearrangement(group_db %>% select(-input_file), file=group_fn)
    cat("- [",group_db_label,"](",file.path("repertoires",basename(group_fn)),")\n", sep="")
}
```

```{r log, echo=FALSE}
cat("START> ClonePass", file=log_fn, append=F)
for (i in 1:nrow(output_groups)) {
    cat(paste0("\nFILE",i,"> ",basename(output_groups[['input_file']][i])), file=log_fn, append=T)
    cat(paste0("\nOUTPUT",i,"> ",basename(output_files[i])), file=log_fn, append=T)
    cat(paste0("\nPASS",i,"> ",output_sizes[i]), file=log_fn, append=T)
    cat(paste0("\nFAIL",i,"> ",input_sizes[i]-output_sizes[i]), file=log_fn, append=T)    
}
```

```{r, child=c('versions.Rmd')}
```